---
title: "Nowcasting a reporting triangle: an end-to-end pipeline"
author: "Richard Aubrey White"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nowcasting a reporting triangle: an end-to-end pipeline}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

`csalert` provides a small, draw-parallel surveillance engine built around three
S3 formats:

- **`csfmt_reporting_triangle_v3`** — the reference-week x reporting-week input
  (who was reported *when*).
- **`csfmt_ensemble_v3`** — `$data` plus, per measure, a matrix of Monte-Carlo
  `$draws` (rows = weeks, columns = simulations). An analysis stage adds columns
  to the draws, so uncertainty propagates without a second pass.
- a **quantile collapse** of those draws (optionally *healed* into
  `cstidy::csfmt_rts_data_v3` for the usual plots and tables).

This vignette runs one synthetic series through four stages:

1. **Nowcast** — complete the right-truncated recent weeks.
2. **Validation** — replay the method against what was known in the past and
   score it.
3. **Reporting completion** — read the reporting delay off the triangle itself.
4. **Short-term trend** — estimate the recent slope of the completed series, per
   draw.

One triangle carries all four, so the stages compose rather than each opening a
fresh demo dataset. When you set up a *new* indicator, run stage 3 first: it is
what supports the choice of `max_delay` that the other three then use.

```{r}
library(data.table)
library(csalert)
```

## A synthetic reporting triangle

Real surveillance data arrives with a delay: a case with reference week `W` may
only be *reported* in week `W`, `W+1`, `W+2`, and so on. The generator below
simulates 70 reference weeks of a seasonal indicator, with a **reporting
speed-up built in at the start of ISO year 2024** — the same system, reporting
faster. Stage 3 recovers that change from the triangle alone.

`set.seed()` is called *inside* the generator, not beside it. A seed set outside
a function that is called more than once leaves the second call running on a
different stream, and the resulting Monte-Carlo noise reads as signal.

```{r}
sim_reports <- function(first_ref = "2023-01",
                        n_weeks = 70L,
                        delay_slow = c(0.45, 0.30, 0.15, 0.07, 0.03),
                        delay_fast = c(0.70, 0.20, 0.06, 0.03, 0.01),
                        seed = 1L) {
  set.seed(seed)                                    # pinned inside, see above
  weeks <- cstime::dates_by_isoyearweek$isoyearweek
  i0 <- match(first_ref, weeks)
  rows <- lapply(seq_len(n_weeks), function(w) {
    ref <- weeks[i0 + w - 1L]
    p <- if (cstime::isoyearweek_to_isoyear_n(ref) <= 2023) delay_slow else delay_fast
    n <- rpois(1, 60 + 25 * sin(2 * pi * w / 52))   # a seasonal signal
    data.table(
      isoyearweek_reference = ref,
      isoyearweek_reporting = weeks[i0 + w - 1L + sample(0:4, n, TRUE, p)]
    )
  })
  # "today" is the last reference week: drop what has not been reported yet, so
  # the most recent weeks are still incomplete -- the problem a nowcast solves
  rbindlist(rows)[isoyearweek_reporting <= weeks[i0 + n_weeks - 1L]]
}

reports <- sim_reports()
triangle_long <- reports[, .(numerator = .N),
                         by = .(isoyearweek_reference, isoyearweek_reporting)]
triangle_long[, `:=`(indicator = "example", location = "nation",
                     age = "total", sex = "total")]
head(triangle_long, 4)
```

Wrap it as a `csfmt_reporting_triangle_v3`. The as-of boundary is read from the
data — it is the newest reporting week present, not the system clock.

```{r}
tri <- csfmt_reporting_triangle_v3(
  triangle_long,
  id_cols       = c("indicator", "location", "age", "sex"),
  reference_col = "isoyearweek_reference",
  reporting_col = "isoyearweek_reporting",
  value_col     = "numerator"
)
c(as_of = attr(tri, "as_of"),
  newest_reporting_week = max(triangle_long$isoyearweek_reporting))
```

`max_delay` is the delay horizon used by stages 1 to 3 below. Stage 3 shows how
to choose it from the data; 5 weeks is the right answer for this series.

```{r}
max_delay <- 5L
```

## 1. Nowcast

**Estimand.** For each reference week, the total that will have been reported at
delays `0 .. max_delay - 1` — that is, by the end of ISO week
`reference_week + max_delay - 1`. This is a horizon-capped total, **not** the
eventual total: anything reported later than `max_delay - 1` weeks is outside the
estimand and no amount of nowcasting recovers it. Stage 3 is how you check that
the horizon is wide enough for the difference to be small. For a settled week the
quantity is already observed; for the most recent weeks it is not, and the
nowcast is a predictive distribution over it.

`nowcast_quasipoisson_v1()` is a discriminative (regression) engine. For each
number of weeks a reference week has been observed it fits, on the settled weeks,
a quasipoisson regression with an identity link of the settled total on the
counts reported so far:

```
total ~ n[delay 0] + n[delay 1] + ... + n[delay h]
```

R's default intercept is present in that formula and is fitted. There is no
per-week magnitude parameter, so the recent weeks do not each carry their own
noisy level. Draws combine the fit's parameter uncertainty with a
dispersion-matched negative binomial. `delay_window` (default 26 weeks) restricts
training to the settled weeks within roughly that span, so the partial-to-total
mapping can follow a reporting regime that changes — as this series' does.

```{r}
set.seed(2)
ens <- nowcast_quasipoisson_v1(tri, max_delay = max_delay, n_sim = 500)
ens
```

Collapse the draws to a quantile summary. `original` is the count reported so
far; the settled weeks sit exactly on it, the recent weeks are completed above
it.

```{r}
q <- ens_collapse(ens, probs = c(0.05, 0.5, 0.95))
tail(q[, .(isoyearweek, original,
           lo  = numerator_nowcasted_q05x0,
           med = numerator_nowcasted_q50x0,
           hi  = numerator_nowcasted_q95x0)], 8)
```

The settled weeks are pinned to their observed total, so they have no band at
all. Only the incomplete weeks at the right-hand edge carry one:

```{r}
show <- tail(q, 30)
xs   <- seq_len(nrow(show))
plot(xs, show$numerator_nowcasted_q50x0, type = "n",
     ylim = range(show$numerator_nowcasted_q05x0, show$numerator_nowcasted_q95x0),
     xlab = "week (last 30)", ylab = "weekly total (nowcast)")
polygon(c(xs, rev(xs)),
        c(show$numerator_nowcasted_q05x0, rev(show$numerator_nowcasted_q95x0)),
        col = grDevices::adjustcolor("steelblue", 0.25), border = NA)
lines(xs, show$numerator_nowcasted_q50x0, lwd = 2, col = "steelblue4")
```

Keep the reference weeks: the later stages index off them.

```{r}
weeks     <- cstime::dates_by_isoyearweek$isoyearweek
ref_weeks <- q$isoyearweek
```

## 2. Validation: replay the method against the past

Because the triangle records *when* every count arrived, you can reconstruct what
was known at any past week and replay the engine against it, without keeping a
second dated extract.

That reconstruction is exact only for an **append-only** reporting system: one
where a count, once filed, keeps its original reporting week forever. Real
systems also issue retrospective corrections, delete records and reclassify
cases, and none of those leave a trace in the current triangle — a case
reclassified last month looks as though it was always classified that way.
Where such revisions matter, replay understates how much the published numbers
actually moved, and a dated archive of extracts is the only way to measure it.

`nowcast_censor()` does the rewind. It returns a `csfmt_reporting_triangle_v3`
with every cell reported after the given week dropped, and its as-of boundary
moved back:

```{r}
past <- nowcast_censor(tri, as_of = ref_weeks[length(ref_weeks) - 8L])
c(now  = attr(tri, "as_of"),  then      = attr(past, "as_of"),
  rows_now = nrow(tri),       rows_then = nrow(past))
```

`nowcast_truth()` supplies the target. It returns a two-column data.table
(`reference`, `truth`) of each reference week's total summed over delays
`0 .. max_delay - 1`, keeping only the weeks old enough for that total to be
settled. The newest weeks are absent by design: they have no truth to be scored
against yet.

```{r}
truth <- nowcast_truth(tri, max_delay = max_delay)
tail(truth, 3)
c(reference_weeks = length(ref_weeks), settled = nrow(truth))
```

A *method* is any function `f(triangle) -> csfmt_ensemble_v3` with its own
parameters baked in. That one-argument contract is what lets engines with
different signatures be replayed and compared through the same harness.

```{r}
method_qp <- function(x) nowcast_quasipoisson_v1(x, max_delay = max_delay, n_sim = 500)
as_of_weeks <- tail(ref_weeks, 30)
```

`nowcast_backtest()` runs the replay and returns the raw scored quantiles: one
long row per `reference` x `as_of` x `horizon` x `quantile_level`, with the
predicted value. `horizon` is weeks between the reference week and the as-of
week, so horizon 0 is the current, least-observed week.

```{r}
bt <- nowcast_backtest(
  tri, method_qp,
  max_delay   = max_delay,
  as_of_weeks = as_of_weeks,
  horizons    = 0:3,
  probs       = c(0.05, 0.25, 0.5, 0.75, 0.95),
  seed        = 1
)
head(bt, 4)
nrow(bt)
```

`nowcast_evaluate_v1()` wraps that replay and scores it, joining each forecast to
its settled truth. It returns one row per group (default: per horizon) per
method:

- `n` — the number of scored forecasts behind that row.
- `coverage_50`, `coverage_90` — the **measured share of settled truths that fell
  inside the nominal 50% and 90% central intervals** on this replay. These are
  sample quantities, not a property of the construction.
- `median_signed`, `median_abs`, `q05`, `q95`, `p_gt_25`, `p_gt_50` — the
  revision of the published median relative to the settled truth, as a fraction
  of the truth. `median_signed` is the **median signed relative revision** across
  the replayed forecasts, not a mean, so it is not the bias in the usual
  expected-error sense. `p_gt_25` and `p_gt_50` are **empirical exceedance
  proportions** — the share of replayed forecasts whose absolute relative
  revision exceeded 0.25 and 0.50 — not probabilities of anything.

Passing a **named list** of methods replays every method over the same reference
and as-of weeks, so the comparison is paired **by forecast unit**. `seed` makes
each method's own run reproducible; it does not by itself create common random
numbers across methods, which would need the algorithms to consume compatible
variates. Here it cannot: `nowcast_passthrough_to_ensemble_v1()` draws no random
numbers at all. Racing against it — it does no completion and just republishes
the counts reported so far — makes the numbers readable:

```{r}
ev <- nowcast_evaluate_v1(
  tri,
  methods = list(
    quasipoisson = method_qp,
    passthrough  = function(x) nowcast_passthrough_to_ensemble_v1(x, max_delay = max_delay)
  ),
  max_delay   = max_delay,
  as_of_weeks = as_of_weeks,
  horizons    = 0:3,
  seed        = 1
)
ev[, .(method, horizon, n, coverage_50, coverage_90, median_signed, median_abs)]
```

Read that table as a measurement on **this** sample and no further. Each row
rests on 26 to 29 scored weeks of one synthetic series, which is far too few to
characterise either engine: a coverage estimate from 26 scored weeks has a
standard error of roughly 0.06 at 0.9 and 0.10 at 0.5, before any dependence
between overlapping windows is allowed for.

What the comparison does show is the *shape* of the problem. What construction
guarantees for the passthrough is only this: each of its forecasts is **no
greater** than the settled truth, because a count that is still arriving cannot
exceed its own total. Every individual revision is therefore zero or negative.
That alone does not force a strictly negative *median* at a given horizon — if
more than half the weeks were already complete, the median would be exactly zero
— and it does not order the horizons. The negative `median_signed` at all four
horizons, and horizon 0 being the worst, are findings on this sample, not
consequences of the construction.

Its `coverage_50` and `coverage_90` are equal because a single draw gives it no
interval at all — the "interval" is a point, so it covers the truth only when the
republished count already equals it, which at horizon 3 happens on the weeks
where nothing arrived at delay 4.

`nowcast_estimate_calibration_v1()` reports the same per-horizon coverage
alongside an interval-scaling factor learned from the replay. Its documentation
is explicit that this is an empirical rescaling and not split conformal, so the
factor carries no finite-sample coverage guarantee.

## 3. Reporting completion: how fast does the data actually arrive?

### What `pct_delayD` counts

**`pct_delay0` is the share of a reference week's cases that were reported during
that same ISO week.** Delay 0 is the reference week itself, not the week after
it. In general:

> `pct_delayD` is the pooled share of a reference week's cases reported by the
> end of ISO week `reference_week + D`.

The columns are indexed by **delay**, 0-based, so the index in the name is the
delay it reports. There are `max_delay` of them and the highest is
`pct_delay<max_delay - 1>`, matching the triangle's own delay axis. Each is the
delay ECDF read at one delay, with no interpolation. `mean_delay` is on the same
axis and in whole weeks, so a week whose cases all arrive at delay 0 has
`mean_delay` 0, not 1.

A triangle with a known answer settles it. Below, each reference week generates
exactly 50 at delay 0, 30 at delay 1 and 20 at delay 2, right-truncated at the
newest reporting week the way real data is:

```{r}
i <- match("2023-01", weeks)
pin <- data.table(
  isoyearweek_reference = weeks[i + rep(0:29, each = 3)],
  isoyearweek_reporting = weeks[i + rep(0:29, each = 3) + rep(0:2, 30)],
  numerator = rep(c(50, 30, 20), 30),
  indicator = "pinned", location = "nation", age = "total", sex = "total"
)
pin <- pin[isoyearweek_reporting <= weeks[i + 29]]
tri_pin <- csfmt_reporting_triangle_v3(
  pin, id_cols = c("indicator", "location", "age", "sex")
)
reporting_completion_v1(tri_pin, max_delay = 3)[
  , .(n_settled, mean_delay, complete_by_md, pct_delay0, pct_delay1, pct_delay2)]
```

`pct_delay0` is 50, not 80: it is the delay-0 share, the reports that arrived in
the reference week itself. `pct_delay1` is 80 — cumulative through delay 1.
`mean_delay` is `0*0.50 + 1*0.30 + 2*0.20 = 0.70`.

**Coming from an older script?** These columns used to be named `pct_w1`,
`pct_w2`, ..., counting weeks-observed from 1, so no number in a column name ever
equalled the delay it stood for. Your `pct_w1` is now `pct_delay0`, `pct_w2` is
`pct_delay1`, and so on. The old names are gone rather than redefined, so old
code errors on a missing column instead of quietly returning a different week.

`n_settled` is 28, not 30. Age eligibility is the first filter: a reference week
is eligible once `as_of_week - reference_week >= max_delay - 1`, which excludes
the two most recent of those 30. A second filter then drops any eligible week
whose total within the horizon is zero, and `n_settled` counts what survives
both. Here no week is empty, so the age rule alone accounts for the number — and
the same holds on the working triangle:

```{r}
age <- match(attr(tri, "as_of"), weeks) - match(ref_weeks, weeks)
completion <- reporting_completion_v1(tri, max_delay = max_delay)
c(age_eligible = sum(age >= max_delay - 1L), reported = completion$n_settled)
```

They agree only because this series has no zero-count weeks. On an indicator
with quiet weeks — a rare pathogen, a small stratum — `n_settled` will be the
smaller of the two, and a period slice with fewer than three surviving weeks is
dropped from the output entirely.

### A worked reference week

Take reference week **2023-07**. `cstime` gives its calendar dates, and the two
weeks after it:

```{r}
cstime::dates_by_isoyearweek[
  isoyearweek %in% c("2023-07", "2023-08", "2023-09"),
  .(isoyearweek, isoyear, mon, thu, sun)
]
```

For a case whose reference week is 2023-07:

| statistic | covers reports up to | which is |
|---|---|---|
| `pct_delay0` | end of ISO week 2023-07 | Sunday 2023-02-19 |
| `pct_delay1` | end of ISO week 2023-08 | Sunday 2023-02-26 |
| `pct_delay2` | end of ISO week 2023-09 | Sunday 2023-03-05 |

So `pct_delay0` is a statement about the seven days from Monday 2023-02-13, and
`pct_delay1` about the **14** days from that same Monday — not the seven days of
week 2023-08 on their own. The columns are cumulative.

### Which day of the week you run it on

**The day of the week does not change the delay arithmetic at all.** Delay is
computed from the two ISO-week labels only, so every day of a week carries the
same label and lands in the same delay bucket:

```{r}
days <- seq(as.Date("2023-02-13"), as.Date("2023-02-19"), by = "day")
data.table(date = days, weekday = weekdays(days),
           isoyearweek = cstime::date_to_isoyearweek_c(days))
```

Both runs put a report filed that week at delay 0 for reference week 2023-07, at
delay 1 for 2023-06, and so on. Nothing in the pipeline reads the system clock:
the as-of boundary comes from the newest reporting week *present in the data*.

That last phrase carries a condition worth stating. `as_of` is
`max(reporting_week)`, so a Monday and a Friday run see the same
`as_of = "2023-07"` **only if the Monday extract already contains at least one
report filed in week 2023-07**. If it contains none — a plausible Monday morning
on a slow indicator — `as_of` silently falls back to `"2023-06"`, every week's
age shifts by one, and one more reference week is treated as settled. That is a
different analysis, not a smaller one, and nothing in the output announces it.

What the day *does* change is **how much of the current week's reporting has
landed**. On Monday almost none of it has; in this illustration all of it is in
by Sunday, which assumes the extract is taken after the week closes and that the
system files everything within the week it belongs to. Neither is guaranteed in
general. In a Monday
extract every cell whose reporting week is the current week is still filling —
the delay-0 cell of the current reference week most visibly, but also the delay-1
cell of last week, the delay-2 cell of the week before, and so on.

That matters for the completion table in one narrow place, and it is worth being
precise about which. Thin the current week's reports down to 15%, as a Monday
extract would see them, and rebuild:

```{r}
set.seed(5)
monday <- reports[isoyearweek_reporting != attr(tri, "as_of") | runif(.N) < 0.15]
tl_mon <- monday[, .(numerator = .N),
                 by = .(isoyearweek_reference, isoyearweek_reporting)]
tl_mon[, `:=`(indicator = "example", location = "nation",
              age = "total", sex = "total")]
tri_mon <- csfmt_reporting_triangle_v3(
  tl_mon, id_cols = c("indicator", "location", "age", "sex")
)

rbind(
  cbind(extract = "Sunday (week complete)",
        reporting_completion_v1(tri, max_delay = max_delay)[
          , .(n_settled, mean_delay, pct_delay0, pct_delay1, pct_delay2, pct_delay3)]),
  cbind(extract = "Monday (15% of the week in)",
        reporting_completion_v1(tri_mon, max_delay = max_delay)[
          , .(n_settled, mean_delay, pct_delay0, pct_delay1, pct_delay2, pct_delay3)])
)
```

The pooled curve barely moves, and `n_settled` is identical, because the current
reference week is never in the settled set. For any `max_delay` of 2 or more its
age is 0, which is below the `max_delay - 1` threshold, so it contributes nothing
to `pct_delayD` whichever day you run on. (A `max_delay` of 1 leaves a single delay
bucket, no completion to measure, and no useful summary; use 2 or more.)

One reference week *is* affected, and it is the newest settled one. Its last
delay cell — delay `max_delay - 1` — is being reported during the current week,
so a mid-week extract sees only part of it. Count the weeks whose settled total
moved, rather than assuming it is one:

```{r}
cmp <- merge(nowcast_truth(tri, max_delay), nowcast_truth(tri_mon, max_delay),
             by = "reference", suffixes = c("_sunday", "_monday"))
cmp[(.N - 2):.N]
cmp[truth_sunday != truth_monday]
```

So: **on this triangle** the day of the week moves the pooled delay curve by
about a tenth of a percentage point. That is not a general result. One settled
week's cell is being diluted into a pool of 66, and the effect scales with how
much weight that single week carries — with three qualifying weeks, or with one
recent week far larger than the rest, the same mechanism is material. Measure it
on your own series before deciding it is negligible.

The newest settled week's *total* is a different matter, and there a mid-week
extract genuinely undercounts. That total is what `nowcast_truth()` scores a
backtest against, so run the backtest off an end-of-week extract, or accept that
the newest scored week enters low.

### "As of today", for the weeks on screen

With `as_of` = `r attr(tri, "as_of")` and `max_delay` = `r max_delay`, the
reference weeks split three ways:

```{r}
data.table(
  isoyearweek = ref_weeks,
  weeks_observed = age + 1L,
  status = fifelse(age >= max_delay - 1L, "settled",
           fifelse(age > 0L, "still filling", "current week"))
)[, .N, keyby = status]
```

Four weeks are not settled: three still filling, plus the current week, which has
only its delay-0 reports. Those are the weeks the nowcast in stage 1 completed,
and the weeks the completion table declines to learn from. The completed weeks
are the ones whose nowcast rises above the count reported so far:

```{r}
q[numerator_nowcasted_q95x0 > original, isoyearweek]
```

### Reporting drift: `period = "year"`

`period = "year"` returns **one row per qualifying ISO year**, and it is how you
see a reporting system speeding up or slowing down. One pooled curve cannot: it
averages the regimes together and describes neither.

"Qualifying" is load-bearing. A period slice needs at least three settled weeks
with a non-zero within-horizon total, or it is dropped from the result with no
warning and no placeholder row. A year at the edge of the series — the one your
data starts or ends in — is the usual casualty, so a missing year means too few
weeks, never zero delay.

```{r}
reporting_completion_v1(tri, max_delay = max_delay, period = "year")[
  , .(period, n_settled, mean_delay, pct_delay0, pct_delay1, pct_delay2, pct_delay3)]
```

That is the change built into `sim_reports()`, recovered from the triangle: in
2023 about 47% of a week's cases were in by the end of the reference week, in
2024 about 68%, and `mean_delay` roughly halved. The pooled row above reports
`pct_delay0` around 52 — a number that was never true of either year.

The stratification is by **ISO year**, not calendar year, and the ISO year of a
week is the calendar year of its Thursday. That decides which year owns a
boundary week:

```{r}
cstime::dates_by_isoyearweek[
  isoyearweek %in% c("2022-52", "2023-01"), .(isoyearweek, isoyear, mon, thu, sun)
]
```

ISO week 2022-52 has its Thursday on 2022-12-29, so it belongs to ISO year 2022
even though its Sunday, 2023-01-01, is a calendar-2023 date. ISO week 2023-01 has
its Thursday on 2023-01-05 and belongs to 2023.

`period = "month"` slices finer and localises *when* a change happened. It uses
the same Thursday rule to decide which calendar month owns a week that straddles
two:

```{r}
tail(reporting_completion_v1(tri, max_delay = max_delay, period = "month")[
  , .(period, n_settled, mean_delay, pct_delay0, pct_delay1)], 6)
```

The step lands between 2023-12 and 2024-01. Note the small `n_settled` per month
— four or five weeks — so a single month's row is noisy; read the sequence, not
one row.

### Every number here is conditional on `max_delay`

This is the trap, and it is structural rather than a tuning subtlety.
`reporting_completion_v1()` works from a triangle that has already had every cell
with delay `>= max_delay` discarded. So the denominator is the total that arrived
*within the horizon*, not the eventual total. Two consequences follow, and they
hold whatever the real reporting tail looks like:

- `complete_by_md` is the last cumulative fraction of that same truncated total,
  so it is 1.
- the last column, `pct_delay<max_delay - 1>`, is that fraction as a percentage,
  so it is 100.

Neither can detect reporting that dribbles in past the horizon. Rather than
assert that, check it — here across `max_delay` 2 through 8 on this triangle,
reading the last column by name each time:

```{r}
sens <- rbindlist(lapply(2:8, function(md) {
  r <- reporting_completion_v1(tri, max_delay = md)
  data.table(max_delay = md, n_settled = r$n_settled, mean_delay = r$mean_delay,
             complete_by_md = r$complete_by_md,
             last_col = paste0("pct_delay", md - 1L),
             last_pct = r[[paste0("pct_delay", md - 1L)]],
             pct_delay0 = r$pct_delay0, pct_delay1 = r$pct_delay1,
             pct_delay2 = r$pct_delay2)
}), fill = TRUE)
sens
c(complete_by_md_always_1 = all(sens$complete_by_md == 1),
  last_pct_always_100     = all(sens$last_pct == 100))
```

The same holds inside every period slice, which is where it is easiest to
mistake for a finding:

```{r}
per <- rbindlist(lapply(c("all", "year", "month"), function(p) {
  r <- reporting_completion_v1(tri, max_delay = max_delay, period = p)
  data.table(period_arg = p, rows = nrow(r),
             complete_by_md_all_1 = all(r$complete_by_md == 1),
             pct_delay4_all_100 = all(r$pct_delay4 == 100))
}))
per
```

**The diagnostic that does work is the `max_delay` sweep itself.** Read the
`sens` table above down its rows: `mean_delay` rises from 0.32 at `max_delay` 2
to 0.82 at 5, then moves only to 0.84 at 8. `pct_delay0` falls from 67.5 to 52.1
and then drifts to 51.0. That flattening is what *supports* `max_delay <- 5L`
here — it is a sensitivity analysis, not a proof. A `mean_delay` that kept
climbing would be clear evidence the tail was still being cut off; a plateau is
weaker evidence in the other direction, because a genuinely sparse tail and a
shifting settled-week composition both flatten the curve too.

Two cautions on reading that sweep:

- The short horizons are not merely imprecise, they are biased optimistic.
  `pct_delay0` at `max_delay` 2 reads 67.5% because it is conditioning on the cases
  that arrived within two weeks — a smaller denominator, so a larger share.
- Not all of the residual movement past 5 is about the tail. `n_settled` falls
  from 66 to 63 across those rows, because a longer horizon settles fewer weeks.
  This series reports faster in 2024, so dropping its three newest weeks lowers
  the pooled `pct_delay0` on composition alone. Compare rows at equal `n_settled`, or
  read `period = "year"` instead, before calling a small drift a tail.

`sim_reports()` emits no delay beyond 4, so on *this* triangle a horizon of 5
truncates nothing and the flattening really is exact — but only because we can
read the generator. On a real series that check is unavailable: widen until
`mean_delay` stops moving, then treat the remaining tail as bounded by what a
still-wider horizon would have shown, not as zero.

## 4. Short-term trend on the nowcast

**Estimand.** The OLS slope of the *completed* series over the last
`trend_isoyearweeks` weeks, and that slope as a percentage of the current level.
This is a **descriptive** quantity: a summary of the six numbers in the window,
not an estimate of a latent growth parameter. Read per draw, it inherits exactly
the uncertainty the nowcast put into those six numbers, and no other.

Running the trend on the nowcast rather than on the reported counts is the point
of the pipeline: the reported counts turn down at the right-hand edge simply
because the reports have not arrived, and a trend fitted to them reports a fall
that is an artefact of reporting.

The method returns a slope, a growth rate and the fraction of draws whose slope
is positive. It does **not** return an increasing / not-increasing
classification; if you want one, you choose the cut-off on that fraction
yourself. (The `cstidy::csfmt_rts_data_v1` method of `short_term_trend()` does
return a status factor — a different estimator, discussed at the end of this
section.)

The ensemble method fits, independently down every draw column, a closed-form OLS
straight line through that draw's last `trend_isoyearweeks` nowcasted counts. It
adds two draw matrices and one point column:

```{r}
ens_trend <- short_term_trend(ens, measure = "numerator_nowcasted",
                              trend_isoyearweeks = 6)
names(ens_trend$draws)
setdiff(names(ens_trend$data), names(ens$data))
```

- `..._trend_beta1` — the OLS slope, in counts per week, per draw.
- `..._trend_gr` — the growth rate `100 * slope / level`, per draw, in percent of
  the current level per week.
- `..._trend_increasing_pr` — the share of draws with a positive slope. It is
  already a reduction over the draw axis, so it lives on `$data` and is not
  collapsed.

Collapse as usual:

```{r}
qt <- ens_collapse(ens_trend, probs = c(0.05, 0.5, 0.95))
trend <- qt[, .(isoyearweek,
                gr    = round(numerator_nowcasted_trend_gr_q50x0, 2),
                gr_lo = round(numerator_nowcasted_trend_gr_q05x0, 2),
                gr_hi = round(numerator_nowcasted_trend_gr_q95x0, 2),
                p_increasing = numerator_nowcasted_trend_increasing_pr)]
tail(trend, 10)
```

The last four rows are the nowcast weeks. Their point estimates are negative —
the seasonal wave in this series is past its peak — but the 5-95% bands cross
zero, so on this series and this window the recent direction is not resolved.
That is the honest reading of a trend fitted to weeks that are still filling.

### The settled weeks have no interval, and that is the right answer

Look at the rows above the nowcast weeks: `gr_lo` and `gr_hi` equal `gr`, and
`p_increasing` is exactly 0 or 1. That is not a display artefact and it is not a
defect. Those windows lie entirely inside the settled weeks, where every draw is
the same observed total. The six counts are *known*, so their OLS slope is a
known number, and the descriptive estimand has nothing left to be uncertain
about. `p_increasing` collapses to a **sign indicator** on that one number —
which is what a share-of-draws becomes when all the draws agree. It is not a
sign test, and no inference is being performed. Measured rather than asserted:

```{r}
settled_weeks <- truth$reference
tr_settled <- trend[isoyearweek %in% settled_weeks & !is.na(gr_lo)]
data.table(
  settled_rows_with_a_full_window = nrow(tr_settled),
  all_intervals_degenerate = all(tr_settled$gr_lo == tr_settled$gr_hi),
  p_increasing_values = paste(sort(unique(tr_settled$p_increasing)), collapse = ", ")
)
```

### `propagate_slope_error = TRUE` changes the estimand

`short_term_trend()` has an argument that gives the settled weeks an interval.
It is worth understanding what it actually does, because it is not "adding the
uncertainty the default forgot".

`propagate_slope_error = TRUE` perturbs each draw's slope by `se * t`, with `se`
the OLS standard error of that window and `t` on `trend_isoyearweeks - 2` degrees
of freedom. That quantity only means something if you stop asking the descriptive
question and start asking a **model-based** one: *treat the six weekly counts as
noisy observations around a latent straight line, and report uncertainty about
that line's slope.* Switching to that question buys an interval, and commits you
to three assumptions:

1. the underlying weekly mean is **linear** across the window;
2. the deviations are **independent** across weeks;
3. they are **homoskedastic**, with a `t` reference distribution.

None of the three holds for the series in this vignette. `sim_reports()` draws
counts from a Poisson whose mean is a **sinusoid**, so the window mean is
curved, not linear, and the variance grows with the level rather than staying
constant. So the interval below is a model-based perturbation under assumptions
the data generator violates — informative about the sensitivity of the slope,
not a valid confidence statement about it.

Use the default when you want to describe what the completed series did. Reach
for this one only when a latent linear trend is genuinely the thing you mean, and
say so when you publish the interval:

```{r}
set.seed(3)
ens_trend2 <- short_term_trend(ens, measure = "numerator_nowcasted",
                               trend_isoyearweeks = 6,
                               propagate_slope_error = TRUE)
qt2 <- ens_collapse(ens_trend2, probs = c(0.05, 0.5, 0.95))
trend2 <- qt2[, .(isoyearweek,
                  gr    = round(numerator_nowcasted_trend_gr_q50x0, 2),
                  gr_lo = round(numerator_nowcasted_trend_gr_q05x0, 2),
                  gr_hi = round(numerator_nowcasted_trend_gr_q95x0, 2),
                  p_increasing = numerator_nowcasted_trend_increasing_pr)]
tail(trend2, 10)
c(degenerate_intervals_now = sum(trend2$gr_lo == trend2$gr_hi, na.rm = TRUE))
```

Every row now has an interval, and `p_increasing` moves off 0 and 1 — but note
what that costs: the settled weeks, whose counts are known exactly, are now
reported as uncertain. That is the change of estimand made visible, not new
information.

One practical trap on top of the assumptions. The degrees of freedom are
`trend_isoyearweeks - 2`, which is 4 at the width of 6 used here. At the
function's own default width of 3 it is 1 — a Cauchy — and the growth-rate
quantiles then have no finite variance and very heavy tails. Widen the window
before enabling this.

This ensemble method is not the same estimator as `short_term_trend()` on a
`cstidy::csfmt_rts_data_v1`. That one fits a quasi-Poisson log-link model over a
moving window and returns a factor status column (`increasing` /
`notincreasing`), following Benedetti (2019); see
`vignette("short_term_trend", package = "csalert")`. The ensemble method here is
an OLS slope on the count scale, chosen because it is a fixed linear filter and
so can be applied down 500 draw columns at once. Use the ensemble method when the
input carries draws, and the `csfmt_rts_data_v1` method when it does not.

## The naming grammar

Measure columns are built from structured components rather than ad-hoc string
pasting, so downstream code routes on parsed parts instead of hard-coded names:

```
<measure>[_vs_<denom>][_<role>][_<q-coord> | _prob_<level>][_pr<per>][<suffix>]
```

```{r}
csfmt_var("numerator", role = "nowcasted", q = 0.5)
csfmt_parse("numerator_nowcasted_q50x0")
```

`csfmt_parse()` is **not** the inverse of `csfmt_var()`. It strips coordinates
right to left and matches a role against a fixed vocabulary, so it cannot tell
which of several role-looking segments was the role. The package's own rate name
is a case where it gets the denominator wrong:

```{r}
nm <- csfmt_var("numerator_nowcasted", denom = "denominator_nowcasted", per = 100)
nm
csfmt_parse(nm)$denom
```

The denominator's own `_nowcasted` was consumed as the role, so `"denominator"`
comes back instead of `"denominator_nowcasted"`. Treat the parse as reliable for
single-role names such as `numerator_nowcasted_q50x0`, and check the result when
a measure or denominator itself ends in a role word.

`q_label()` and `q_value()` map a probability to and from its column label, and
are likewise not a clean inverse pair: `q_label()` holds one decimal-percent
digit and exactly two integer-percent digits, so a finer probability is rounded
and `p = 1` produces a three-digit label that `q_value()` cannot read back.

```{r}
q_value(q_label(0.0125))   # rounded to the label grid
q_label(1)
q_value(q_label(1))        # NA: three integer digits do not parse
```

## Where next

- Add a rate with `ens_add_rate()` (a nowcasted numerator over a nowcasted
  denominator), MEM intensity thresholds with `mem_thresholds_v1()`, or
  exceedance detection with `signal_detection_hlm()`. All three write a new
  matrix into `$draws` before the collapse, so they compose with stage 4 above.
- `ens_collapse(heal = TRUE)` returns a `cstidy::csfmt_rts_data_v3` for the
  standard plotting and table helpers.
- `reporting_completion_trend_v1()` wraps stage 3's year and month slices into
  one table with a `scope` column.
- `nowcast_estimate_calibration_v1()` turns a long replay into a per-horizon
  interval-width scaling factor, as a diagnostic on an engine.
- For the quasi-Poisson trend estimator on `cstidy` data, including the county
  map example, see `vignette("short_term_trend", package = "csalert")`.
