---
title: "Nowcasting a reporting triangle: an end-to-end pipeline"
author: "Richard Aubrey White"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nowcasting a reporting triangle: an end-to-end pipeline}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

`csalert` provides a small, draw-parallel surveillance engine built around three
S3 formats:

- **`csfmt_reporting_triangle_v3`** — the reference-week x reporting-week input
  (who was reported *when*).
- **`csfmt_ensemble_v3`** — `$data` plus, per measure, a matrix of Monte-Carlo
  `$draws` (rows = weeks, columns = simulations). An analysis stage adds columns
  to the draws, so uncertainty propagates without a second pass.
- a **quantile collapse** of those draws (optionally *healed* into
  `cstidy::csfmt_rts_data_v3` for the usual plots and tables).

## The ensemble is the analysis substrate

One rule explains the shape of everything below.

> **Every analytical stage takes a `csfmt_ensemble_v3` and returns a
> `csfmt_ensemble_v3`.** `ens_collapse()` is **terminal**: it reduces the draws
> to quantiles, and a collapsed table is output, never input.

So the canonical pipeline is a single chain through the ensemble:

```
csfmt_reporting_triangle_v3
  -> nowcast_quasipoisson_v1  OR  nowcast_passthrough_to_ensemble_v1
  -> [ens_add_rate]              # when a denominator exists
  -> [short_term_trend]          # growth rate + P(increasing)
  -> [mem_thresholds_v1]         # MEM intensity
  -> [signal_detection_hlm]      # per-draw exceedance
  -> ens_collapse(heal = TRUE)   # ensemble -> quantiles -> csfmt_rts_data_v3
```

The bracketed stages are optional and order-flexible among themselves; each one
adds draw columns and hands the ensemble on. What you cannot do is collapse and
then carry on: a collapsed table holds quantiles, not draws, so feeding it back
into `short_term_trend()` cannot recover a per-draw trend or a P(increasing).
That is by design, not a gap — the draws are the uncertainty, and the collapse
is where you spend them.

This vignette runs one synthetic series through the whole chain:

1. **Nowcast** — complete the right-truncated recent weeks.
2. **Validation** — replay the method against what was known in the past.
3. **Reporting completion** — read the reporting delay off the triangle itself.
4. **Rate** — a nowcasted numerator over a nowcasted denominator, per draw.
5. **Short-term trend** — the recent slope of that rate, per draw.
6. **MEM intensity** — seasonal intensity thresholds, classified per draw.
7. **Signal detection** — historical-limits exceedance, per draw.
8. **Collapse** — quantiles, healed into `cstidy::csfmt_rts_data_v3`.

Stages 4 to 8 are the ensemble chain above. Stages 2 and 3 are not: they take the
**triangle**, not the ensemble, because they ask about the reporting process
rather than about the completed counts. That asymmetry is worth remembering —
diagnostics of *how data arrives* read the triangle; analyses of *what the data
says* read the ensemble.

When you set up a *new* indicator, run stage 3 first: it is what supports the
choice of `max_delay` that everything else then uses.

```{r}
library(data.table)
library(csalert)
```

## A synthetic reporting triangle

Real surveillance data arrives with a delay: a case with reference week `W` may
only be *reported* in week `W`, `W+1`, `W+2`, and so on. The generator below
simulates a laboratory indicator from ISO week 2019-35 to 2024-18 — five winter
seasons, which is what stages 6 and 7 need — with a **reporting speed-up built in
at the start of ISO year 2024**. Stage 3 recovers that change from the triangle
alone.

It emits a **numerator and a denominator**: tests taken, and how many were
positive. Simulating positives as a binomial *conditional on the tests* keeps the
numerator a subset of the denominator, which is what makes their ratio a
proportion. Two independent Poissons would not be. Each test carries one
reporting delay, so a result and its test land in the same triangle cell.

`set.seed()` is called *inside* the generator, not beside it. A seed set outside
a function that is called more than once leaves the second call running on a
different stream, and the resulting Monte-Carlo noise reads as signal.

```{r}
sim_reports <- function(first_ref = "2019-35",
                        last_ref = "2024-18",
                        delay_slow = c(0.45, 0.30, 0.15, 0.07, 0.03),
                        delay_fast = c(0.70, 0.20, 0.06, 0.03, 0.01),
                        seed = 1L) {
  set.seed(seed)                                    # pinned inside, see above
  weeks <- cstime::dates_by_isoyearweek$isoyearweek
  i0 <- match(first_ref, weeks)
  i1 <- match(last_ref, weeks)
  rows <- lapply(i0:i1, function(ri) {
    ref <- weeks[ri]
    s <- cos(2 * pi * (cstime::isoyearweek_to_isoweek_n(ref) - 5) / 52)  # winter peak
    n_tests <- rpois(1, 400 + 120 * s)
    positive <- rbinom(n_tests, 1, stats::plogis(-2.2 + 1.1 * s))        # subset of tests
    p <- if (cstime::isoyearweek_to_isoyear_n(ref) <= 2023) delay_slow else delay_fast
    data.table(
      isoyearweek_reference = ref,
      isoyearweek_reporting = weeks[ri + sample(0:4, n_tests, TRUE, p)],
      positive = positive
    )
  })
  # "today" is the last reference week: drop what has not been reported yet, so
  # the most recent weeks are still incomplete -- the problem a nowcast solves
  rbindlist(rows)[isoyearweek_reporting <= weeks[i1]]
}

reports <- sim_reports()
triangle_long <- reports[, .(numerator = sum(positive), denominator = .N),
                         by = .(isoyearweek_reference, isoyearweek_reporting)]
triangle_long[, `:=`(indicator_tag = "lab_flu", location_code = "nation",
                     age = "total", sex = "total")]
head(triangle_long, 4)
c(cells = nrow(triangle_long),
  numerator_never_exceeds_denominator =
    all(triangle_long$numerator <= triangle_long$denominator))
```

Wrap it as a `csfmt_reporting_triangle_v3`. The as-of boundary is read from the
data — it is the newest reporting week present, not the system clock.

```{r}
tri <- csfmt_reporting_triangle_v3(
  triangle_long,
  id_cols       = c("indicator_tag", "location_code", "age", "sex"),
  reference_col = "isoyearweek_reference",
  reporting_col = "isoyearweek_reporting",
  value_col     = "numerator"
)
c(as_of = attr(tri, "as_of"),
  newest_reporting_week = max(triangle_long$isoyearweek_reporting))
```

`max_delay` is the delay horizon used by stages 1 to 3 below. Stage 3 shows how
to choose it from the data; 5 weeks is the right answer for this series.

```{r}
max_delay <- 5L
```

## 1. Nowcast

**Estimand.** For each reference week, the total that will have been reported at
delays `0 .. max_delay - 1` — that is, by the end of ISO week
`reference_week + max_delay - 1`. This is a horizon-capped total, **not** the
eventual total: anything reported later than `max_delay - 1` weeks is outside the
estimand and no amount of nowcasting recovers it. Stage 3 is how you check that
the horizon is wide enough for the difference to be small. For a settled week the
quantity is already observed; for the most recent weeks it is not, and the
nowcast is a predictive distribution over it.

`nowcast_quasipoisson_v1()` is a discriminative (regression) engine. For each
number of weeks a reference week has been observed it fits, on the settled weeks,
a quasipoisson regression with an identity link of the settled total on the
counts reported so far:

```
total ~ n[delay 0] + n[delay 1] + ... + n[delay h]
```

R's default intercept is present in that formula and is fitted. There is no
per-week magnitude parameter, so the recent weeks do not each carry their own
noisy level. Draws combine the fit's parameter uncertainty with a
dispersion-matched negative binomial. `delay_window` (default 26 weeks) restricts
training to the settled weeks within roughly that span, so the partial-to-total
mapping can follow a reporting regime that changes — as this series' does.

`denominator_col` nowcasts a second measure alongside the numerator, on the same
draw axis. Stage 4 needs it: a rate is only coherent if both its parts are
completed in the same Monte-Carlo world.

```{r}
set.seed(2)
ens <- nowcast_quasipoisson_v1(tri, max_delay = max_delay, n_sim = 500,
                               denominator_col = "denominator")
ens
```

Collapse the draws to a quantile summary. `original` is the count reported so
far; the settled weeks sit exactly on it, the recent weeks are completed above
it.

```{r}
q <- ens_collapse(ens, probs = c(0.05, 0.5, 0.95))
tail(q[, .(isoyearweek, original,
           lo  = numerator_nowcasted_q05x0,
           med = numerator_nowcasted_q50x0,
           hi  = numerator_nowcasted_q95x0)], 8)
```

The settled weeks are pinned to their observed total, so they have no band at
all. Only the incomplete weeks at the right-hand edge carry one:

```{r}
show <- tail(q, 30)
xs   <- seq_len(nrow(show))
plot(xs, show$numerator_nowcasted_q50x0, type = "n",
     ylim = range(show$numerator_nowcasted_q05x0, show$numerator_nowcasted_q95x0),
     xlab = "week (last 30)", ylab = "weekly total (nowcast)")
polygon(c(xs, rev(xs)),
        c(show$numerator_nowcasted_q05x0, rev(show$numerator_nowcasted_q95x0)),
        col = grDevices::adjustcolor("steelblue", 0.25), border = NA)
lines(xs, show$numerator_nowcasted_q50x0, lwd = 2, col = "steelblue4")
```

Keep the reference weeks: the later stages index off them.

```{r}
weeks     <- cstime::dates_by_isoyearweek$isoyearweek
ref_weeks <- q$isoyearweek
```

## 2. Validation: replay the method against the past

This stage and the next take the **triangle**, not the ensemble: they ask how the
data arrives, which is a property of the reporting process and is destroyed the
moment the delay axis is collapsed into weekly totals.

Because the triangle records *when* every count arrived, you can reconstruct what
was known at any past week and replay the engine against it, without keeping a
second dated extract.

That reconstruction is exact only for an **append-only** reporting system: one
where a count, once filed, keeps its original reporting week forever. Real
systems also issue retrospective corrections, delete records and reclassify
cases, and none of those leave a trace in the current triangle — a case
reclassified last month looks as though it was always classified that way.
Where such revisions matter, replay understates how much the published numbers
actually moved, and a dated archive of extracts is the only way to measure it.

`nowcast_censor()` does the rewind. It returns a `csfmt_reporting_triangle_v3`
with every cell reported after the given week dropped, and its as-of boundary
moved back:

```{r}
past <- nowcast_censor(tri, as_of = ref_weeks[length(ref_weeks) - 8L])
c(now  = attr(tri, "as_of"),  then      = attr(past, "as_of"),
  rows_now = nrow(tri),       rows_then = nrow(past))
```

`nowcast_truth()` supplies the target. It returns a two-column data.table
(`reference`, `truth`) of each reference week's total summed over delays
`0 .. max_delay - 1`, keeping only the weeks old enough for that total to be
settled. The newest weeks are absent by design: they have no truth to be scored
against yet.

```{r}
truth <- nowcast_truth(tri, max_delay = max_delay)
tail(truth, 3)
c(reference_weeks = length(ref_weeks), settled = nrow(truth))
```

A *method* is any function `f(triangle) -> csfmt_ensemble_v3` with its own
parameters baked in. That one-argument contract is what lets engines with
different signatures be replayed and compared through the same harness.

```{r}
method_qp <- function(x) nowcast_quasipoisson_v1(x, max_delay = max_delay, n_sim = 500)
as_of_weeks <- tail(ref_weeks, 30)
```

`nowcast_backtest()` runs the replay and returns the raw scored quantiles: one
long row per `reference` x `as_of` x `horizon` x `quantile_level`, with the
predicted value. `horizon` is weeks between the reference week and the as-of
week, so horizon 0 is the current, least-observed week.

```{r}
bt <- nowcast_backtest(
  tri, method_qp,
  max_delay   = max_delay,
  as_of_weeks = as_of_weeks,
  horizons    = 0:3,
  probs       = c(0.05, 0.25, 0.5, 0.75, 0.95),
  seed        = 1
)
head(bt, 4)
nrow(bt)
```

`nowcast_evaluate_v1()` wraps that replay and scores it, joining each forecast to
its settled truth. It returns one row per group (default: per horizon) per
method:

- `n` — the number of scored forecasts behind that row.
- `coverage_50`, `coverage_90` — the **measured share of settled truths that fell
  inside the nominal 50% and 90% central intervals** on this replay. These are
  sample quantities, not a property of the construction.
- `median_signed`, `median_abs`, `q05`, `q95`, `p_gt_25`, `p_gt_50` — the
  revision of the published median relative to the settled truth, as a fraction
  of the truth. `median_signed` is the **median signed relative revision** across
  the replayed forecasts, not a mean, so it is not the bias in the usual
  expected-error sense. `p_gt_25` and `p_gt_50` are **empirical exceedance
  proportions** — the share of replayed forecasts whose absolute relative
  revision exceeded 0.25 and 0.50 — not probabilities of anything.

Passing a **named list** of methods replays every method over the same reference
and as-of weeks, so the comparison is paired **by forecast unit**. `seed` makes
each method's own run reproducible; it does not by itself create common random
numbers across methods, which would need the algorithms to consume compatible
variates. Here it cannot: `nowcast_passthrough_to_ensemble_v1()` draws no random
numbers at all. Racing against it — it does no completion and just republishes
the counts reported so far — makes the numbers readable:

```{r}
ev <- nowcast_evaluate_v1(
  tri,
  methods = list(
    quasipoisson = method_qp,
    passthrough  = function(x) nowcast_passthrough_to_ensemble_v1(x, max_delay = max_delay)
  ),
  max_delay   = max_delay,
  as_of_weeks = as_of_weeks,
  horizons    = 0:3,
  seed        = 1
)
ev[, .(method, horizon, n, coverage_50, coverage_90, median_signed, median_abs)]
```

Read that table as a measurement on **this** sample and no further. Each row
rests on 26 to 29 scored weeks of one synthetic series, which is far too few to
characterise either engine: a coverage estimate from 26 scored weeks has a
standard error of roughly 0.06 at 0.9 and 0.10 at 0.5, before any dependence
between overlapping windows is allowed for.

What the comparison does show is the *shape* of the problem. What construction
guarantees for the passthrough is only this: each of its forecasts is **no
greater** than the settled truth, because a count that is still arriving cannot
exceed its own total. Every individual revision is therefore zero or negative.
That alone does not force a strictly negative *median* at a given horizon — if
more than half the weeks were already complete, the median would be exactly zero
— and it does not order the horizons. The negative `median_signed` at all four
horizons, and horizon 0 being the worst, are findings on this sample, not
consequences of the construction.

Its `coverage_50` and `coverage_90` are equal because a single draw gives it no
interval at all — the "interval" is a point, so it covers the truth only when the
republished count already equals it, which at horizon 3 happens on the weeks
where nothing arrived at delay 4.

### The engine's horizon-0 row is worth stopping on

The quasipoisson engine's horizon-0 coverage is well below nominal, and its
`median_signed` is *positive* — it is over-completing. That is not a random dip.
The replay window above ends at the as-of week, so it straddles the reporting
speed-up this series has at the start of ISO 2024, and the engine trains on the
26 preceding settled weeks. Learn the completion factors of a slow regime, apply
them to partial counts from a fast one, and you scale up counts that were already
nearly complete.

The contrast is visible if the replay is restricted to as-of weeks that sit
entirely inside the slow regime:

```{r}
slow_weeks <- tail(ref_weeks[cstime::isoyearweek_to_isoyear_n(ref_weeks) == 2023], 30)
c(from = slow_weeks[1], to = slow_weeks[length(slow_weeks)])
nowcast_evaluate_v1(tri, method_qp, max_delay = max_delay,
                    as_of_weeks = slow_weeks, horizons = 0:3, seed = 1)[
  , .(horizon, n, coverage_50, coverage_90, median_signed, median_abs)]
```

Horizon 0 recovers, and the bias changes sign. Read that as a demonstration of
the mechanism, not as a measured effect size: the two windows also differ in
which weeks and which part of the season they cover, and 26 to 30 scored weeks
cannot separate those from the regime change. The general lesson is the one
`delay_window` exists for — **a nowcast engine is only as current as the
reporting behaviour it was trained on**, and stage 3 is how you find out when
that behaviour moved.

`nowcast_estimate_calibration_v1()` reports the same per-horizon coverage
alongside an interval-scaling factor learned from the replay. Its documentation
is explicit that this is an empirical rescaling and not split conformal, so the
factor carries no finite-sample coverage guarantee.

## 3. Reporting completion: how fast does the data actually arrive?

### What `pct_delayD` counts

**`pct_delay0` is the share of a reference week's cases that were reported during
that same ISO week.** Delay 0 is the reference week itself, not the week after
it. In general:

> `pct_delayD` is the pooled share of a reference week's cases reported by the
> end of ISO week `reference_week + D`.

The columns are indexed by **delay**, 0-based, so the index in the name is the
delay it reports. There are `max_delay` of them and the highest is
`pct_delay<max_delay - 1>`, matching the triangle's own delay axis. Each is the
delay ECDF read at one delay, with no interpolation. `mean_delay` is on the same
axis and in whole weeks, so a week whose cases all arrive at delay 0 has
`mean_delay` 0, not 1.

A triangle with a known answer settles it. Below, each reference week generates
exactly 50 at delay 0, 30 at delay 1 and 20 at delay 2, right-truncated at the
newest reporting week the way real data is:

```{r}
i <- match("2023-01", weeks)
pin <- data.table(
  isoyearweek_reference = weeks[i + rep(0:29, each = 3)],
  isoyearweek_reporting = weeks[i + rep(0:29, each = 3) + rep(0:2, 30)],
  numerator = rep(c(50, 30, 20), 30),
  indicator = "pinned", location = "nation", age = "total", sex = "total"
)
pin <- pin[isoyearweek_reporting <= weeks[i + 29]]
tri_pin <- csfmt_reporting_triangle_v3(
  pin, id_cols = c("indicator", "location", "age", "sex")
)
reporting_completion_v1(tri_pin, max_delay = 3)[
  , .(n_settled, mean_delay, complete_by_md, pct_delay0, pct_delay1, pct_delay2)]
```

`pct_delay0` is 50, not 80: it is the delay-0 share, the reports that arrived in
the reference week itself. `pct_delay1` is 80 — cumulative through delay 1.
`mean_delay` is `0*0.50 + 1*0.30 + 2*0.20 = 0.70`.

**Coming from an older script?** These columns used to be named `pct_w1`,
`pct_w2`, ..., counting weeks-observed from 1, so no number in a column name ever
equalled the delay it stood for. Your `pct_w1` is now `pct_delay0`, `pct_w2` is
`pct_delay1`, and so on. The old names are gone rather than redefined, so old
code errors on a missing column instead of quietly returning a different week.

`n_settled` is 28, not 30. Age eligibility is the first filter: a reference week
is eligible once `as_of_week - reference_week >= max_delay - 1`, which excludes
the two most recent of those 30. A second filter then drops any eligible week
whose total within the horizon is zero, and `n_settled` counts what survives
both. Here no week is empty, so the age rule alone accounts for the number — and
the same holds on the working triangle:

```{r}
age <- match(attr(tri, "as_of"), weeks) - match(ref_weeks, weeks)
completion <- reporting_completion_v1(tri, max_delay = max_delay)
c(age_eligible = sum(age >= max_delay - 1L), reported = completion$n_settled)
```

They agree only because this series has no zero-count weeks. On an indicator
with quiet weeks — a rare pathogen, a small stratum — `n_settled` will be the
smaller of the two, and a period slice with fewer than three surviving weeks is
dropped from the output entirely.

### A worked reference week

Take reference week **2023-07**. `cstime` gives its calendar dates, and the two
weeks after it:

```{r}
cstime::dates_by_isoyearweek[
  isoyearweek %in% c("2023-07", "2023-08", "2023-09"),
  .(isoyearweek, isoyear, mon, thu, sun)
]
```

For a case whose reference week is 2023-07:

| statistic | covers reports up to | which is |
|---|---|---|
| `pct_delay0` | end of ISO week 2023-07 | Sunday 2023-02-19 |
| `pct_delay1` | end of ISO week 2023-08 | Sunday 2023-02-26 |
| `pct_delay2` | end of ISO week 2023-09 | Sunday 2023-03-05 |

So `pct_delay0` is a statement about the seven days from Monday 2023-02-13, and
`pct_delay1` about the **14** days from that same Monday — not the seven days of
week 2023-08 on their own. The columns are cumulative.

### Which day of the week you run it on

**The day of the week does not change the delay arithmetic at all.** Delay is
computed from the two ISO-week labels only, so every day of a week carries the
same label and lands in the same delay bucket:

```{r}
days <- seq(as.Date("2023-02-13"), as.Date("2023-02-19"), by = "day")
data.table(date = days, weekday = weekdays(days),
           isoyearweek = cstime::date_to_isoyearweek_c(days))
```

Both runs put a report filed that week at delay 0 for reference week 2023-07, at
delay 1 for 2023-06, and so on. Nothing in the pipeline reads the system clock:
the as-of boundary comes from the newest reporting week *present in the data*.

That last phrase carries a condition worth stating. `as_of` is
`max(reporting_week)`, so a Monday and a Friday run see the same
`as_of = "2023-07"` **only if the Monday extract already contains at least one
report filed in week 2023-07**. If it contains none — a plausible Monday morning
on a slow indicator — `as_of` silently falls back to `"2023-06"`, every week's
age shifts by one, and one more reference week is treated as settled. That is a
different analysis, not a smaller one, and nothing in the output announces it.

What the day *does* change is **how much of the current week's reporting has
landed**. On Monday almost none of it has; in this illustration all of it is in
by Sunday, which assumes the extract is taken after the week closes and that the
system files everything within the week it belongs to. Neither is guaranteed in
general. In a Monday
extract every cell whose reporting week is the current week is still filling —
the delay-0 cell of the current reference week most visibly, but also the delay-1
cell of last week, the delay-2 cell of the week before, and so on.

That matters for the completion table in one narrow place, and it is worth being
precise about which. Thin the current week's reports down to 15%, as a Monday
extract would see them, and rebuild:

```{r}
set.seed(5)
monday <- reports[isoyearweek_reporting != attr(tri, "as_of") | runif(.N) < 0.15]
tl_mon <- monday[, .(numerator = sum(positive), denominator = .N),
                 by = .(isoyearweek_reference, isoyearweek_reporting)]
tl_mon[, `:=`(indicator_tag = "lab_flu", location_code = "nation",
              age = "total", sex = "total")]
tri_mon <- csfmt_reporting_triangle_v3(
  tl_mon, id_cols = c("indicator_tag", "location_code", "age", "sex")
)

rbind(
  cbind(extract = "Sunday (week complete)",
        reporting_completion_v1(tri, max_delay = max_delay)[
          , .(n_settled, mean_delay, pct_delay0, pct_delay1, pct_delay2, pct_delay3)]),
  cbind(extract = "Monday (15% of the week in)",
        reporting_completion_v1(tri_mon, max_delay = max_delay)[
          , .(n_settled, mean_delay, pct_delay0, pct_delay1, pct_delay2, pct_delay3)])
)
```

The pooled curve barely moves, and `n_settled` is identical, because the current
reference week is never in the settled set. For any `max_delay` of 2 or more its
age is 0, which is below the `max_delay - 1` threshold, so it contributes nothing
to `pct_delayD` whichever day you run on. (A `max_delay` of 1 leaves a single delay
bucket, no completion to measure, and no useful summary; use 2 or more.)

At most **one** reference week can have its settled total affected: the newest
settled one, whose last delay cell — delay `max_delay - 1` — is still being
reported during the current week. Every older week finished reporting earlier;
every newer week is not settled. Count the weeks that actually moved, rather than
assuming it is one:

```{r}
cmp <- merge(nowcast_truth(tri, max_delay), nowcast_truth(tri_mon, max_delay),
             by = "reference", suffixes = c("_sunday", "_monday"))
cmp[(.N - 2):.N]
cmp[truth_sunday != truth_monday]
```

On this series, none did — and the reason is worth seeing, because it bounds the
whole effect. The only cell at risk is the delay-4 numerator of the newest
settled week, and in the fast 2024 regime delay 4 carries 1% of a week's tests:

```{r}
newest_settled <- weeks[match(attr(tri, "as_of"), weeks) - (max_delay - 1L)]
triangle_long[isoyearweek_reference == newest_settled &
              isoyearweek_reporting == attr(tri, "as_of"),
              .(isoyearweek_reference, isoyearweek_reporting, numerator, denominator)]
```

Five tests, none of them positive. Thinning zero positives leaves zero, so the
numerator's settled total cannot move. **The weekday effect is bounded by the
mass sitting in the last delay bucket** — widen `max_delay` past where reporting
actually finishes and that bucket empties, which is why it vanishes here and did
not vanish on a series with a heavier tail.

**None of this generalises, and the two reasons it vanishes here are both about
scale.** The at-risk cell is one week's last delay bucket, diluted into a pool of
241 settled weeks; and that bucket is nearly empty because `max_delay` is wide
enough. Shrink the series, shrink the horizon, or give the indicator a heavier
reporting tail, and the same mechanism becomes material — a period slice with the
minimum three qualifying weeks gives that one cell a third of the weight.

The general statement is the conditional one: a mid-week extract undercounts the
newest settled week's total by whatever share of its last delay bucket has not
arrived. That total is what `nowcast_truth()` scores a backtest against, so run
the backtest off an end-of-week extract, or measure the gap on your own series
rather than assuming it is as small as it is here.

### "As of today", for the weeks on screen

With `as_of` = `r attr(tri, "as_of")` and `max_delay` = `r max_delay`, the
reference weeks split three ways:

```{r}
data.table(
  isoyearweek = ref_weeks,
  weeks_observed = age + 1L,
  status = fifelse(age >= max_delay - 1L, "settled",
           fifelse(age > 0L, "still filling", "current week"))
)[, .N, keyby = status]
```

Four weeks are not settled: three still filling, plus the current week, which has
only its delay-0 reports. Those are the weeks the nowcast in stage 1 completed,
and the weeks the completion table declines to learn from. The completed weeks
are the ones whose nowcast rises above the count reported so far:

```{r}
q[numerator_nowcasted_q95x0 > original, isoyearweek]
```

### Reporting drift: `period = "year"`

`period = "year"` returns **one row per qualifying ISO year**, and it is how you
see a reporting system speeding up or slowing down. One pooled curve cannot: it
averages the regimes together and describes neither.

"Qualifying" is load-bearing. A period slice needs at least three settled weeks
with a non-zero within-horizon total, or it is dropped from the result with no
warning and no placeholder row. A year at the edge of the series — the one your
data starts or ends in — is the usual casualty, so a missing year means too few
weeks, never zero delay.

```{r}
reporting_completion_v1(tri, max_delay = max_delay, period = "year")[
  , .(period, n_settled, mean_delay, pct_delay0, pct_delay1, pct_delay2, pct_delay3)]
```

That is the change built into `sim_reports()`, recovered from the triangle. Five
consecutive ISO years sit within a couple of points of each other — `pct_delay0`
around 43 to 45, `mean_delay` around 0.95 — and then 2024 steps to about 69% and
0.45. A flat run followed by a step is what a genuine regime change looks like;
a single year out of line with its neighbours is usually noise.

The pooled row from the previous table reports `pct_delay0` of 47.3, which is
close to the five slow years only because they outnumber the fast one five to
one. It is not a compromise between the regimes; it is the old regime with a
little contamination, and it will drift year by year as 2024 accumulates weeks.

The stratification is by **ISO year**, not calendar year, and the ISO year of a
week is the calendar year of its Thursday. That decides which year owns a
boundary week:

```{r}
cstime::dates_by_isoyearweek[
  isoyearweek %in% c("2022-52", "2023-01"), .(isoyearweek, isoyear, mon, thu, sun)
]
```

ISO week 2022-52 has its Thursday on 2022-12-29, so it belongs to ISO year 2022
even though its Sunday, 2023-01-01, is a calendar-2023 date. ISO week 2023-01 has
its Thursday on 2023-01-05 and belongs to 2023.

`period = "month"` slices finer and localises *when* a change happened. It uses
the same Thursday rule to decide which calendar month owns a week that straddles
two:

```{r}
tail(reporting_completion_v1(tri, max_delay = max_delay, period = "month")[
  , .(period, n_settled, mean_delay, pct_delay0, pct_delay1)], 6)
```

The step lands between 2023-12 and 2024-01. Note the small `n_settled` per month
— four or five weeks — so a single month's row is noisy; read the sequence, not
one row.

### Every number here is conditional on `max_delay`

This is the trap, and it is structural rather than a tuning subtlety.
`reporting_completion_v1()` works from a triangle that has already had every cell
with delay `>= max_delay` discarded. So the denominator is the total that arrived
*within the horizon*, not the eventual total. Two consequences follow, and they
hold whatever the real reporting tail looks like:

- `complete_by_md` is the last cumulative fraction of that same truncated total,
  so it is 1.
- the last column, `pct_delay<max_delay - 1>`, is that fraction as a percentage,
  so it is 100.

Neither can detect reporting that dribbles in past the horizon. Rather than
assert that, check it — here across `max_delay` 2 through 8 on this triangle,
reading the last column by name each time:

```{r}
sens <- rbindlist(lapply(2:8, function(md) {
  r <- reporting_completion_v1(tri, max_delay = md)
  data.table(max_delay = md, n_settled = r$n_settled, mean_delay = r$mean_delay,
             complete_by_md = r$complete_by_md,
             last_col = paste0("pct_delay", md - 1L),
             last_pct = r[[paste0("pct_delay", md - 1L)]],
             pct_delay0 = r$pct_delay0, pct_delay1 = r$pct_delay1,
             pct_delay2 = r$pct_delay2)
}), fill = TRUE)
sens
c(complete_by_md_always_1 = all(sens$complete_by_md == 1),
  last_pct_always_100     = all(sens$last_pct == 100))
```

The same holds inside every period slice, which is where it is easiest to
mistake for a finding:

```{r}
per <- rbindlist(lapply(c("all", "year", "month"), function(p) {
  r <- reporting_completion_v1(tri, max_delay = max_delay, period = p)
  data.table(period_arg = p, rows = nrow(r),
             complete_by_md_all_1 = all(r$complete_by_md == 1),
             pct_delay4_all_100 = all(r$pct_delay4 == 100))
}))
per
```

**The diagnostic that does work is the `max_delay` sweep itself.** Read the
`sens` table above down its rows: `mean_delay` rises from 0.38 at `max_delay` 2
to 0.88 at 5, then moves only to 0.89 at 8. `pct_delay0` falls from 61.8 to 47.3
and then drifts to 46.8. That flattening is what *supports* `max_delay <- 5L`
here — it is a sensitivity analysis, not a proof. A `mean_delay` that kept
climbing would be clear evidence the tail was still being cut off; a plateau is
weaker evidence in the other direction, because a genuinely sparse tail and a
shifting settled-week composition both flatten the curve too.

Two cautions on reading that sweep:

- The short horizons are not merely imprecise, they are biased optimistic.
  `pct_delay0` at `max_delay` 2 reads 61.8% because it is conditioning on the
  cases that arrived within two weeks — a smaller denominator, so a larger share.
- Not all of the residual movement past 5 is about the tail. `n_settled` falls
  from 244 to 238 across those rows, because a longer horizon settles fewer
  weeks, and the weeks it drops are the newest — which on this series are the
  fast-reporting ones. That pulls the pooled `pct_delay0` down slightly on
  composition alone. Compare rows at equal `n_settled`, or read
  `period = "year"` instead, before calling a small drift a tail.

`sim_reports()` emits no delay beyond 4, so on *this* triangle a horizon of 5
truncates nothing and the flattening really is exact — but only because we can
read the generator. On a real series that check is unavailable: widen until
`mean_delay` stops moving, then treat the remaining tail as bounded by what a
still-wider horizon would have shown, not as zero.

## 4. Rate: a nowcasted numerator over a nowcasted denominator

From here on every stage takes the ensemble and returns the ensemble.

**Estimand.** The percentage of tests that were positive, per reference week —
computed **per draw**, so it carries the uncertainty of both the numerator and
the denominator. Collapsing first and dividing the medians would throw that away
and give a ratio no draw ever produced.

Because draws are index-aligned across measures (column `j` is the same
Monte-Carlo world for every measure), the division is element-wise:

```{r}
ens <- ens_add_rate(ens,
                    numerator   = "numerator_nowcasted",
                    denominator = "denominator_nowcasted",
                    per         = 100)
names(ens$draws)
```

The new measure's name is built by the grammar, not pasted, so build it the same
way rather than typing it out:

```{r}
rate <- csfmt_var("numerator_nowcasted", denom = "denominator_nowcasted", per = 100)
rate
```

```{r}
qr <- ens_collapse(ens, probs = c(0.05, 0.5, 0.95))
tail(qr[, .(isoyearweek,
            positives = numerator_nowcasted_q50x0,
            tests     = denominator_nowcasted_q50x0,
            pct_lo    = round(get(csfmt_var(rate, q = 0.05)), 2),
            pct       = round(get(csfmt_var(rate, q = 0.50)), 2),
            pct_hi    = round(get(csfmt_var(rate, q = 0.95)), 2))], 6)
```

Two guards are worth knowing about. A denominator of zero gives `NA`, not a
fabricated 0% that would read as a real drop. And because the numerator is a
subset of the denominator, the rate is capped at `per`; a draw that violated that
would warn rather than silently exceed 100%. No warning appeared above, so no
draw did — but the numerator and denominator here are nowcast **independently**,
and nothing in the engine enforces coherence between them. On a series where the
two are close, expect that warning.

## 5. Short-term trend on the rate

**Estimand.** The OLS slope of the *completed* series over the last
`trend_isoyearweeks` weeks, and that slope as a percentage of the current level.
This is a **descriptive** quantity: a summary of the six numbers in the window,
not an estimate of a latent growth parameter. Read per draw, it inherits exactly
the uncertainty the nowcast put into those six numbers, and no other.

Running the trend on the nowcast rather than on the reported counts is the point
of the pipeline: the reported counts turn down at the right-hand edge simply
because the reports have not arrived, and a trend fitted to them reports a fall
that is an artefact of reporting.

The method returns a slope, a growth rate and the fraction of draws whose slope
is positive. It does **not** return an increasing / not-increasing
classification; if you want one, you choose the cut-off on that fraction
yourself. (The `cstidy::csfmt_rts_data_v1` method of `short_term_trend()` does
return a status factor — a different estimator, discussed at the end of this
section.)

The ensemble method fits, independently down every draw column, a closed-form OLS
straight line through that draw's last `trend_isoyearweeks` nowcasted counts. It
adds two draw matrices and one point column:

```{r}
ens_trend <- short_term_trend(ens, measure = rate, trend_isoyearweeks = 6)
setdiff(names(ens_trend$draws), names(ens$draws))
setdiff(names(ens_trend$data), names(ens$data))
```

- `..._trend_beta1` — the OLS slope, in percentage points per week, per draw.
- `..._trend_gr` — the growth rate `100 * slope / level`, per draw, in percent of
  the current level per week.
- `..._trend_increasing_pr` — the share of draws with a positive slope. It is
  already a reduction over the draw axis, so it lives on `$data` and is not
  collapsed.

Collapse as usual:

```{r}
gr <- csfmt_var(rate, role = "trend", suffix = "_gr")
qt <- ens_collapse(ens_trend, probs = c(0.05, 0.5, 0.95))
trend <- qt[, .(isoyearweek,
                gr    = round(get(csfmt_var(gr, q = 0.50)), 2),
                gr_lo = round(get(csfmt_var(gr, q = 0.05)), 2),
                gr_hi = round(get(csfmt_var(gr, q = 0.95)), 2),
                p_increasing = get(csfmt_var(rate, role = "trend",
                                             suffix = "_increasing_pr")))]
tail(trend, 10)
```

The last four rows are the nowcast weeks. Their point estimates are negative and
their 5-95% bands sit entirely below zero, with `p_increasing` at 0 — the series
is on the spring side of its winter peak, and the nowcast uncertainty is not
wide enough to admit a rise. Note what that statement is and is not: it says the
*completed* rate fell over each six-week window, given this nowcast. It does not
say the fall will continue, and it is conditional on the nowcast being right
about the four weeks that are still filling — which stage 2 measured as the
engine's weakest point.

### The settled weeks have no interval, and that is the right answer

Look at the rows above the nowcast weeks: `gr_lo` and `gr_hi` equal `gr`, and
`p_increasing` is exactly 0 or 1. That is not a display artefact and it is not a
defect. Those windows lie entirely inside the settled weeks, where every draw of
the numerator and of the denominator is the same observed total, so every draw of
their ratio is too. The six rates are *known*, so their OLS slope is a known
number, and the descriptive estimand has nothing left to be uncertain
about. `p_increasing` collapses to a **sign indicator** on that one number —
which is what a share-of-draws becomes when all the draws agree. It is not a
sign test, and no inference is being performed. Measured rather than asserted:

```{r}
settled_weeks <- truth$reference
tr_settled <- trend[isoyearweek %in% settled_weeks & !is.na(gr_lo)]
data.table(
  settled_rows_with_a_full_window = nrow(tr_settled),
  all_intervals_degenerate = all(tr_settled$gr_lo == tr_settled$gr_hi),
  p_increasing_values = paste(sort(unique(tr_settled$p_increasing)), collapse = ", ")
)
```

### `propagate_slope_error = TRUE` changes the estimand

`short_term_trend()` has an argument that gives the settled weeks an interval.
It is worth understanding what it actually does, because it is not "adding the
uncertainty the default forgot".

`propagate_slope_error = TRUE` perturbs each draw's slope by `se * t`, with `se`
the OLS standard error of that window and `t` on `trend_isoyearweeks - 2` degrees
of freedom. That quantity only means something if you stop asking the descriptive
question and start asking a **model-based** one: *treat the six weekly counts as
noisy observations around a latent straight line, and report uncertainty about
that line's slope.* Switching to that question buys an interval, and commits you
to three assumptions:

1. the underlying weekly mean is **linear** across the window;
2. the deviations are **independent** across weeks;
3. they are **homoskedastic**, with a `t` reference distribution.

None of the three holds for the series in this vignette. `sim_reports()` draws
tests from a Poisson and positives from a binomial, both with a **sinusoidal**
mean, so the window mean is curved rather than linear and the variance of the
rate changes with the level and with the number of tests. So the interval below
is a model-based perturbation under assumptions the data generator violates —
informative about the sensitivity of the slope, not a valid confidence statement
about it.

Use the default when you want to describe what the completed series did. Reach
for this one only when a latent linear trend is genuinely the thing you mean, and
say so when you publish the interval:

```{r}
set.seed(3)
ens_trend2 <- short_term_trend(ens, measure = rate, trend_isoyearweeks = 6,
                               propagate_slope_error = TRUE)
qt2 <- ens_collapse(ens_trend2, probs = c(0.05, 0.5, 0.95))
trend2 <- qt2[, .(isoyearweek,
                  gr    = round(get(csfmt_var(gr, q = 0.50)), 2),
                  gr_lo = round(get(csfmt_var(gr, q = 0.05)), 2),
                  gr_hi = round(get(csfmt_var(gr, q = 0.95)), 2))]
tail(trend2, 10)
c(degenerate_intervals_now = sum(trend2$gr_lo == trend2$gr_hi, na.rm = TRUE))
```

Every row now has an interval, and `p_increasing` moves off 0 and 1 — but note
what that costs: the settled weeks, whose counts are known exactly, are now
reported as uncertain. That is the change of estimand made visible, not new
information.

One practical trap on top of the assumptions. The degrees of freedom are
`trend_isoyearweeks - 2`, which is 4 at the width of 6 used here. At the
function's own default width of 3 it is 1 — a Cauchy — and the growth-rate
quantiles then have no finite variance and very heavy tails. Widen the window
before enabling this.

This ensemble method is not the same estimator as `short_term_trend()` on a
`cstidy::csfmt_rts_data_v1`. That one fits a quasi-Poisson log-link model over a
moving window and returns a factor status column (`increasing` /
`notincreasing`), following Benedetti (2019); see
`vignette("short_term_trend", package = "csalert")`. The ensemble method here is
an OLS slope, chosen because it is a fixed linear filter and so can be applied
down 500 draw columns at once. **The `csfmt_rts_data_v1` method is the
pre-ensemble architecture and is deprecated**; new work should run
`short_term_trend()` on the ensemble, as above, before `ens_collapse()`.

Carry the trend ensemble forward — the chain continues from here:

```{r}
ens <- ens_trend
```

## 6. MEM intensity thresholds

**Estimand.** Which of five seasonal intensity levels this week's rate falls in,
where the thresholds come from the *same series' own previous seasons* via the
Moving Epidemic Method. Classification happens per draw, so the answer is a
distribution over levels rather than one label.

This is the first stage that needs **history**, and a lot of it:
`mem_thresholds_v1()` needs complete prior seasons to fit a season's thresholds —
`min_seasons` (default 2) is a hard floor, `prefer_seasons` (default 5) is the
depth below which a fit is reported as provisional, and a season needs
`min_weeks_per_season` (default 30) weeks to count as training at all. It is the
reason this vignette's series starts in 2019 rather than last year.

```{r}
ens <- mem_thresholds_v1(ens, measure = rate)
tail(ens$data[, .(isoyearweek, mem_n_seasons, mem_preepidemic,
                  mem_medium, mem_high, mem_veryhigh)], 3)
```

The **message** is not noise to skip past: it says some seasons were fit on fewer
than `prefer_seasons` training seasons, and `mem_n_seasons` records how many each
one actually got. Thresholds fit on two seasons are worth much less than
thresholds fit on five, and this is where you find out which you have.

The classification arrives as per-level probabilities after the collapse:

```{r}
qm <- ens_collapse(ens, probs = 0.5)
pcols <- grep("_status_prob_", names(qm), value = TRUE)
intensity <- qm[, c("isoyearweek", pcols), with = FALSE]
setnames(intensity, pcols, sub(".*_status_prob_", "", pcols))
tail(intensity, 6)
```

Weeks whose season has no thresholds — the early seasons, which have no prior
seasons to learn from — get `NA` for every draw and are not classified at all:

```{r}
status <- ens$draws[[csfmt_var(rate, role = "status")]]
c(weeks = nrow(status),
  weeks_with_no_threshold = sum(apply(status, 1, function(r) all(is.na(r)))))
```

A real influenza series would also want `exclude_seasons` to keep anomalous
seasons — a pandemic year, a season with a data gap — out of the training
baseline. Thresholds are still estimated *for* an excluded season; only the
baseline they are fit on changes.

## 7. Signal detection: historical limits

**Estimand.** Whether this week's rate exceeds what the same calendar week has
looked like in previous years — the historical-limits method. The baseline is the
mean and standard deviation of the same week (plus or minus one) across
`baseline_isoyears` prior years, and the threshold is its 99.5th percentile.
Every draw is compared against that threshold, so the output is an exceedance
*probability*, not a yes/no flag.

The default `baseline_isoyears = 5` needs five full years of history before the
first week can be classified. This series is shorter than that, so it uses three,
which is the kind of trade every new indicator faces:

```{r}
ens <- signal_detection_hlm(ens, measure = rate, baseline_isoyears = 3)
qh <- ens_collapse(ens, probs = 0.5)
hcols <- grep("_hlmstatus_prob_", names(qh), value = TRUE)
signal <- qh[, c("isoyearweek", "hlm_threshold", hcols), with = FALSE]
setnames(signal, hcols, sub(".*_hlmstatus_prob_", "p_", hcols))
tail(signal, 6)
```

`p_high` is the share of draws above the threshold. It is a statement about
*nowcast* uncertainty — how likely the completed rate exceeds the historical
limit — and not a p-value, a posterior probability, or a false-alarm rate. The
baseline itself is estimated from the point history with no uncertainty attached,
so `hlm_threshold` is treated as known.

```{r}
hlm <- ens$draws[[csfmt_var(rate, role = "hlmstatus")]]
c(weeks = nrow(hlm),
  weeks_with_no_baseline = sum(apply(hlm, 1, function(r) all(is.na(r)))))
```

## 8. Collapse: the end of the chain

`ens_collapse()` reduces every draw matrix over the draw axis into quantile
columns, and every ordinal status matrix into per-level probabilities. With
`heal = TRUE` it hands the result to `cstidy::set_csfmt_rts_data_v3()`, which
adds the standard calendar columns and returns a `csfmt_rts_data_v3`.

```{r}
final <- ens_collapse(ens, probs = c(0.05, 0.5, 0.95), heal = TRUE)
class(final)
c(rows = nrow(final), cols = ncol(final))
tail(final[, .(isoyearweek, isoyear, isoweek, season, seasonweek)], 3)
```

**This is where the pipeline ends.** `csfmt_rts_data_v3` is a presentation and
storage format: plots, tables, reports. It is not an analysis substrate, and
there is no route back — the draws are gone, so a per-draw trend, a MEM class
probability or an exceedance probability computed after this point is not
available. Everything that needs draws must happen before the collapse, which is
why stages 4 to 7 are all upstream of it.

On storage, one caveat worth stating plainly: `csdb` currently ships table
validators for `csfmt_rts_data_v1` and `v2` only. There is no `v3` validator yet,
so writing a healed v3 to a database is not something this pipeline can do today.

## The naming grammar

Measure columns are built from structured components rather than ad-hoc string
pasting, so downstream code routes on parsed parts instead of hard-coded names:

```
<measure>[_vs_<denom>][_<role>][_<q-coord> | _prob_<level>][_pr<per>][<suffix>]
```

```{r}
csfmt_var("numerator", role = "nowcasted", q = 0.5)
csfmt_parse("numerator_nowcasted_q50x0")
```

`csfmt_parse()` is **not** the inverse of `csfmt_var()`. It strips coordinates
right to left and matches a role against a fixed vocabulary, so it cannot tell
which of several role-looking segments was the role. The package's own rate name
is a case where it gets the denominator wrong:

```{r}
nm <- csfmt_var("numerator_nowcasted", denom = "denominator_nowcasted", per = 100)
nm
csfmt_parse(nm)$denom
```

The denominator's own `_nowcasted` was consumed as the role, so `"denominator"`
comes back instead of `"denominator_nowcasted"`. Treat the parse as reliable for
single-role names such as `numerator_nowcasted_q50x0`, and check the result when
a measure or denominator itself ends in a role word.

`q_label()` and `q_value()` map a probability to and from its column label, and
are likewise not a clean inverse pair: `q_label()` holds one decimal-percent
digit and exactly two integer-percent digits, so a finer probability is rounded
and `p = 1` produces a three-digit label that `q_value()` cannot read back.

```{r}
q_value(q_label(0.0125))   # rounded to the label grid
q_label(1)
q_value(q_label(1))        # NA: three integer digits do not parse
```

## Where next

- Run the chain on several series at once. Every stage above is written over the
  `time_series_id` axis, so a triangle with many locations or age groups flows
  through unchanged; the seam masking in the trend and the per-series MEM fits
  are already handled.
- `nowcast_passthrough_to_ensemble_v1()` substitutes for the engine at stage 1
  when an indicator should not be nowcast-completed. Everything downstream is
  identical, so an indicator can opt out of nowcasting without opting out of the
  pipeline.
- `reporting_completion_trend_v1()` wraps stage 3's year and month slices into
  one table with a `scope` column.
- `nowcast_estimate_calibration_v1()` turns a long replay into a per-horizon
  interval-width scaling factor, as a diagnostic on an engine.
- For the quasi-Poisson trend estimator on `cstidy` data, including the county
  map example, see `vignette("short_term_trend", package = "csalert")`.
