---
title: "The pipeline: from incomplete counts to published numbers"
author: "Richard Aubrey White"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The pipeline: from incomplete counts to published numbers}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

Surveillance counts arrive late. A case from week `W` can be reported in week
`W`, `W+1`, `W+2` or later. So the newest weeks look lower than they will be.
A trend, an intensity level or an alert measured on them has the same dip.

This vignette takes one synthetic series through csalert. It completes the
weeks that are still arriving, measures how far to trust that completion, and
ends with numbers that you can publish. Three formats carry the run:

- **`csfmt_reporting_triangle_v3`**, the input. It holds one cell per
  reference week and reporting DATE, so it records what was reported *when*.
  A delay is a number of days.
- **`csfmt_ensemble_v3`**, the working format. It holds `$data`, and one
  matrix of Monte-Carlo `$draws` per measure, with rows = weeks and columns =
  draws. Each stage adds draw columns, so the uncertainty travels forward.
- **a quantile collapse** of those draws, the output. It can be *healed* into
  `cstidy::csfmt_rts_data_v3` for the usual plots and tables.

## One rule: every stage takes the ensemble and returns it

> **Every analysis stage takes a `csfmt_ensemble_v3` and returns a
> `csfmt_ensemble_v3`.** `ens_collapse()` is the last stage: it reduces the
> draws to quantiles, and a collapsed table is output, never input.

So the pipeline is one chain through the ensemble:

```
csfmt_reporting_triangle_v3
  -> nowcast_delay_ecdf_v1  OR  nowcast_passthrough_to_ensemble_v1
  -> [ens_add_rate]              # when a denominator exists
  -> [short_term_trend]          # growth rate + P(increasing)
  -> [mem_thresholds_v1]         # MEM intensity
  -> [signal_detection_hlm]      # per-draw exceedance
  -> ens_collapse(heal = TRUE)   # ensemble -> quantiles -> csfmt_rts_data_v3
```

The stages in brackets are optional, and their order is free. Each adds draw
columns and passes the ensemble on. A collapsed table holds quantiles, not
draws, so `short_term_trend()` on it cannot get back a per-draw trend or a
P(increasing). The draws are the uncertainty, and the collapse is where you
spend them.

The run below has nine stages:

1. **Nowcast**: complete the recent weeks that are still being reported.
2. **Validation**: replay the method on what was known in the past.
3. **Reporting completion**: read the reporting delay from the triangle.
4. **Rate**: a nowcast numerator over a nowcast denominator, per draw.
5. **Short-term trend**: the recent slope of that rate, per draw.
6. **MEM intensity**: seasonal intensity thresholds, per draw.
7. **Signal detection**: historical-limits exceedance, per draw.
8. **Collapse**: quantiles, healed into `cstidy::csfmt_rts_data_v3`.
9. **Quality control**: check the input, and compare two runs.

Stages 4 to 8 are the ensemble chain above. Stages 2 and 3 take the
**triangle**, because they ask how the data arrives, not what the completed
counts say. A question about *how the data arrives* reads the triangle, and a
question about *what the data says* reads the ensemble.

For a *new* indicator, run stage 3 first. It supports the choice of
`max_delay_days`, which the other stages then use.

```{r}
library(data.table)
library(csalert)
```

## A synthetic reporting triangle

The generator below simulates a laboratory indicator from ISO week 2019-35 to
2024-18. That span holds five winter seasons, which stages 6 and 7 need.
Reporting **speeds up at the start of ISO year 2024**, and stage 3 finds that
change from the triangle alone.

**The reporting axis is a calendar date, so a delay is a number of days.** The
generator draws a whole-week delay, then a day inside that reporting week. So a
report lands from day 0, the Monday of the reference week, to day 34.

It emits a **numerator and a denominator**: positive tests, and tests taken.
It draws the positives as a binomial *given the tests*, so the numerator is a
subset of the denominator and their ratio is a proportion. Two independent
Poissons would not give one. Each test has one reporting delay, so a result and
its test land in the same triangle cell.

The generator calls `set.seed()` *inside* itself. With the seed outside, a
second call runs on a different random stream, and that Monte-Carlo noise then
reads as signal.

```{r}
sim_reports <- function(first_ref = "2019-35",
                        last_ref = "2024-18",
                        delay_slow = c(0.45, 0.30, 0.15, 0.07, 0.03),
                        delay_fast = c(0.70, 0.20, 0.06, 0.03, 0.01),
                        seed = 1L) {
  set.seed(seed)                                    # pinned inside, see above
  weeks <- cstime::dates_by_isoyearweek$isoyearweek
  monday <- as.Date(cstime::dates_by_isoyearweek$mon)   # each week's own Monday
  i0 <- match(first_ref, weeks)
  i1 <- match(last_ref, weeks)
  rows <- lapply(i0:i1, function(ri) {
    ref <- weeks[ri]
    s <- cos(2 * pi * (cstime::isoyearweek_to_isoweek_n(ref) - 5) / 52)  # winter peak
    n_tests <- rpois(1, 400 + 120 * s)
    positive <- rbinom(n_tests, 1, stats::plogis(-2.2 + 1.1 * s))        # subset of tests
    p <- if (cstime::isoyearweek_to_isoyear_n(ref) <= 2023) delay_slow else delay_fast
    data.table(
      isoyearweek_reference = ref,
      # a whole-week delay, then a day inside that reporting week: delay day
      # 7 * weeks_late + day_of_week, so 0 to 34
      reporting_date = monday[ri] + 7L * sample(0:4, n_tests, TRUE, p) +
        sample(0:6, n_tests, TRUE),
      positive = positive
    )
  })
  # "today" is the last day of the last reference week: drop what has not been
  # reported yet, so the most recent weeks are still incomplete -- the problem a
  # nowcast solves
  rbindlist(rows)[reporting_date <= monday[i1] + 6L]
}

reports <- sim_reports()
triangle_long <- reports[, .(numerator = sum(positive), denominator = .N),
                         by = .(isoyearweek_reference, reporting_date)]
triangle_long[, `:=`(indicator_tag = "lab_flu", location_code = "nation",
                     age = "total", sex = "total")]
head(triangle_long, 4)
c(cells = nrow(triangle_long),
  numerator_never_exceeds_denominator =
    all(triangle_long$numerator <= triangle_long$denominator))
```

Wrap it as a `csfmt_reporting_triangle_v3`. The as-of boundary comes from the
data: it is the newest reporting date, not the system clock. The constructor
stops on a reporting column that is not a `Date`, and on a missing reporting
date.

```{r}
tri <- csfmt_reporting_triangle_v3(
  triangle_long,
  id_cols       = c("indicator_tag", "location_code", "age", "sex"),
  reference_col = "isoyearweek_reference",
  reporting_col = "reporting_date",
  value_col     = "numerator"
)
c(as_of = attr(tri, "as_of"),
  newest_reporting_date = max(triangle_long$reporting_date))
```

`max_delay_days` is the delay horizon of stages 1 to 3, in DAYS. Stage 3 shows
how to choose it. For this series it is 35: delay days 0 to 34, the five weeks
that start on the Monday of the reference week.

```{r}
max_delay_days <- 35L
```

## 1. Nowcast

**Estimand.** For each reference week, the total reported at any delay day from
0 on. A report at delay day `max_delay_days - 1` or later counts in the last
delay column, so it stays in the total. The horizon sets how many delay columns
the nowcast models, not which reports count. For a settled week the total is
observed, except for reports that still arrive after delay day
`max_delay_days - 1`. For the most recent weeks it is not observed, and the
nowcast is a predictive distribution of it. Stage 3 shows how much arrives in
that last column, which is how you check that the horizon is wide enough.

`nowcast_delay_ecdf_v1()` completes each incomplete week from a pool of settled
reference weeks. Take a reference week of age `d` days, so delay days 0 to `d`
are observed. For each pool week `s`, `T_s` is its settled total and `O_s` is
its count by delay day `d`. The draws scale the count of the week by the
quantiles of the completion ratio of the pool:

```
draw[ref] = observed_so_far[ref] * quantile(T_s / O_s)
```

The quantiles sit at `n_sim` evenly spaced probabilities from 0 to 1, in random
order. So the 5 to 95 band is the observed count times the 5% and 95% pool
ratios, to within the spacing of those probabilities. Nothing parametric
is added, and no regression is fitted. A regression on 35 daily delay columns
would have more predictors than settled training rows.

The engine also pools a **daily delay ECDF**, `p(d)`, the share of the counts
of a week that has arrived by delay day `d`. `p(d)` cancels out of every draw.
It only decides whether delay day `d` is completed at all. So the draws are not
built from `observed_so_far / p(d)`, the closed-form maximum likelihood
estimate under `n[ref, d] ~ Poisson(lambda[ref] * p[d])`.

`delay_window`, default 26 WEEKS and the one quantity here that is not days,
limits the pool to the settled weeks of about that span. The completion ratios
can then follow a reporting pattern that changes, as this one does.

`denominator_col` nowcasts a second measure on the same draw axis. Stage 4
needs it: a rate is coherent only if both its parts are completed in the same
Monte-Carlo draw.

```{r}
set.seed(2)
ens <- nowcast_delay_ecdf_v1(tri, max_delay_days = max_delay_days, n_sim = 500,
                             denominator_col = "denominator")
ens
```

Collapse the draws to quantiles. `original` is the count reported so far. A
settled week sits exactly on it, and a recent week is completed above it.

```{r}
q <- ens_collapse(ens, probs = c(0.05, 0.5, 0.95))
tail(q[, .(isoyearweek, original,
           lo  = numerator_nowcasted_q05x0,
           med = numerator_nowcasted_q50x0,
           hi  = numerator_nowcasted_q95x0)], 8)
```

The settled weeks have no band. Only the incomplete weeks at the right-hand
edge carry one:

```{r}
show <- tail(q, 30)
xs   <- seq_len(nrow(show))
plot(xs, show$numerator_nowcasted_q50x0, type = "n",
     ylim = range(show$numerator_nowcasted_q05x0, show$numerator_nowcasted_q95x0),
     xlab = "week (last 30)", ylab = "weekly total (nowcast)")
polygon(c(xs, rev(xs)),
        c(show$numerator_nowcasted_q05x0, rev(show$numerator_nowcasted_q95x0)),
        col = grDevices::adjustcolor("steelblue", 0.25), border = NA)
lines(xs, show$numerator_nowcasted_q50x0, lwd = 2, col = "steelblue4")
```

Keep the reference weeks and their Mondays. Every delay and every age below is
a number of days between two dates, so the later stages need the Mondays, not
the week labels.

```{r}
weeks      <- cstime::dates_by_isoyearweek$isoyearweek
ref_weeks  <- q$isoyearweek
ref_monday <- as.Date(cstime::dates_by_isoyearweek$mon[match(ref_weeks, weeks)])
age        <- as.integer(attr(tri, "as_of") - ref_monday)   # age in DAYS
c(weeks = length(ref_weeks), oldest_age_days = max(age), newest_age_days = min(age))
```

## 2. Validation: replay the method on the past

This stage and the next take the **triangle**, because they ask how the data
arrives. That is a property of the reporting, and a collapse of the delay axis
into weekly totals removes it.

The triangle records the day on which each count arrived. So you can rebuild
what was known on any past date and replay the engine on it, with no second
dated extract.

That rebuild is exact only for an **append-only** reporting system, in which a
count keeps its reporting date for ever. Real systems also correct counts,
delete records and reclassify cases, and none of those leave a trace in the
current triangle. A case reclassified last month looks as though it always had
its new class. Where such revisions matter, the replay understates how much the
published numbers moved. Only a dated archive of extracts can measure that.

`nowcast_censor()` goes back in time. It takes a `Date`, and returns the
triangle without the cells reported after that date, with its as-of boundary
moved back:

```{r}
week_end <- ref_monday + 6L                       # the Sunday that ends each week
past <- nowcast_censor(tri, as_of = week_end[length(week_end) - 8L])
# format() first: c() on a Date and an integer coerces the integer to a Date
c(now  = format(attr(tri, "as_of")), then     = format(attr(past, "as_of")),
  rows_now = format(nrow(tri)),      rows_then = format(nrow(past)))
```

`nowcast_truth()` gives the target: a data.table with `reference` and `truth`,
the total of each reference week over delay days `0` to `max_delay_days - 1`.
It keeps only the weeks old enough for that total to be settled, so the newest
weeks have no truth yet.

```{r}
truth <- nowcast_truth(tri, max_delay_days = max_delay_days)
tail(truth, 3)
c(reference_weeks = length(ref_weeks), settled = nrow(truth))
```

A *method* is any function that takes a triangle and returns a
`csfmt_ensemble_v3`, with its own arguments fixed. That one-argument contract
lets the same code replay and compare engines with different arguments.

`as_of_weeks` MUST be a `Date` vector, and any other class is an error. Replay
as of the Sunday that ends each week, which is what "as of week W" meant when
delays were counted in weeks.

```{r}
method_ecdf <- function(x) {
  nowcast_delay_ecdf_v1(x, max_delay_days = max_delay_days, n_sim = 500)
}
as_of_dates <- tail(week_end, 30)
```

`nowcast_backtest()` runs the replay and returns the quantiles: one row per
`reference`, `as_of`, `horizon` and `quantile_level`, with the `predicted`
value. `horizon` is the number of whole weeks from the reference week to the
as-of date. Horizon 0 is the current week, which is the least observed.

```{r}
bt <- nowcast_backtest(
  tri, method_ecdf,
  max_delay_days = max_delay_days,
  as_of_weeks    = as_of_dates,
  horizons       = 0:3,
  probs          = c(0.05, 0.25, 0.5, 0.75, 0.95),
  seed           = 1
)
head(bt, 4)
nrow(bt)
```

`nowcast_evaluate_v1()` runs the replay and scores each forecast against its
settled truth. It returns one row per horizon and method:

- `n`: the number of scored forecasts in the row.
- `coverage_50`, `coverage_90`: the **measured share of settled truths inside
  the nominal 50% and 90% central intervals** on this replay. These are sample
  quantities, not a property of the method.
- `median_signed`, `median_abs`, `q05`, `q95`, `p_gt_25`, `p_gt_50`: the
  revision of the published median against the settled truth, as a fraction of
  the truth. `median_signed` is the **median signed relative revision** over
  the replayed forecasts. It is a median, not a mean, so it is not the bias in
  the usual expected-error sense. `p_gt_25` and `p_gt_50` are **empirical
  exceedance proportions**: the share of replayed forecasts whose absolute
  relative revision was above 0.25 and 0.50. They are not probabilities of
  anything.

A **named list** of methods replays every method on the same reference and
as-of weeks, so the comparison is paired **by forecast**. `seed` makes the run
of each method reproducible. It does not give common random numbers across
methods, which would need the methods to use their random numbers in the same
way. Here they cannot: `nowcast_passthrough_to_ensemble_v1()` draws no random
numbers at all. It does no completion and republishes the counts reported so
far, which makes the numbers easy to read:

```{r}
ev <- nowcast_evaluate_v1(
  tri,
  methods = list(
    delay_ecdf  = method_ecdf,
    passthrough = function(x) {
      nowcast_passthrough_to_ensemble_v1(x, max_delay_days = max_delay_days)
    }
  ),
  max_delay_days = max_delay_days,
  as_of_weeks    = as_of_dates,
  horizons       = 0:3,
  seed           = 1
)
ev[, .(method, horizon, n, coverage_50, coverage_90, median_signed, median_abs)]
```

Read that table as a measurement on **this** sample only. Each row rests on 26
to 29 scored weeks of one synthetic series, far too few to characterise either
engine. A coverage estimate from 26 scored weeks has a standard error of about
0.06 at 0.9 and 0.10 at 0.5. That is before any allowance for dependence
between overlapping windows.

The comparison does show the *shape* of the problem. For the passthrough,
construction guarantees one thing: each forecast is **no greater** than the
settled truth. A count still arriving cannot exceed its own total, so every
single revision is zero or negative.

That alone does not force a negative *median* at a given horizon. If more than
half the weeks were already complete, the median would be exactly zero. Nor
does it order the horizons. The negative `median_signed` at all four horizons,
and horizon 0 as the worst, are findings on this sample, not results of the
construction.

Its `coverage_50` and `coverage_90` are equal, because a single draw gives no
interval, only a point. The point covers the truth only when the count
reported so far already equals it. At horizon 3 that happens on the weeks where
nothing arrived on delay days 28 to 34.

### The engine completes too much at horizon 0

The `delay_ecdf` engine covers 0.731 of the settled truths at horizon 0 against
a nominal 0.90, and its `median_signed` is *positive*, at 0.389. That is not a
random dip. The replay window ends at the as-of date, so it spans the reporting
speed-up at the start of ISO 2024. `delay_window` trains on the settled weeks of
about the 26 weeks before. So the engine learns the delay curve of the slow
regime, and applies it to the counts of the fast one. It scales up counts that
were already nearly complete.

A replay on as-of weeks inside the slow regime shows the contrast:

```{r}
slow_dates <- tail(week_end[cstime::isoyearweek_to_isoyear_n(ref_weeks) == 2023], 30)
c(from = slow_dates[1], to = slow_dates[length(slow_dates)])
nowcast_evaluate_v1(tri, method_ecdf, max_delay_days = max_delay_days,
                    as_of_weeks = slow_dates, horizons = 0:3, seed = 1)[
  , .(horizon, n, coverage_50, coverage_90, median_signed, median_abs)]
```

Horizon 0 recovers, and the sign of the revision flips. Read that as a
demonstration of the mechanism, not as a measured effect size. The two windows
also differ in their weeks and in their part of the season. 26 to 30 scored
weeks cannot separate those from the regime change. The general lesson is the
reason `delay_window` exists: **a nowcast engine is only as current as the
reporting it was trained on**. Stage 3 shows when that reporting changed.

`nowcast_estimate_calibration_v1()` reports the same coverage per horizon, with
an interval scaling factor learned from the replay. Its help page states that
this is an empirical rescaling and not split conformal, so the factor carries no
finite-sample coverage guarantee.

## 3. Reporting completion: how fast does the data arrive?

### What `pct_delayD` counts

**The delay axis is DAYS.** `pct_delay0` is the share of the cases of a
reference week reported on its own Monday. In general:

> `pct_delayD` is the pooled share of the cases of a reference week reported by
> the end of the day `reference Monday + D`.

The index in the name is the delay day, from 0. There are exactly
`max_delay_days` columns, and the last is `pct_delay<max_delay_days - 1>`. Each
is the delay ECDF read at one day, with no interpolation. `mean_delay` is on the
same axis, in DAYS.

Five of the 35 columns fall on the end of an ISO week, and a weekly reader
wants those: `pct_delay6`, `pct_delay13`, `pct_delay20`, `pct_delay27` and
`pct_delay34`. `pct_delay6` is the share in by the end of the reference week
itself. The other four are the ends of the four weeks after it.

A triangle with a known answer settles it. Each reference week below reports
exactly 50 on delay day 0, 30 on delay day 7 and 20 on delay day 14. The data
stops at the newest reporting date, as real data does:

```{r}
pin_monday <- as.Date("2023-01-02") + 7 * rep(0:29, each = 3)
pin <- data.table(
  isoyearweek_reference = format(pin_monday, "%G-%V"),
  reporting_date = pin_monday + rep(c(0, 7, 14), 30),
  numerator = rep(c(50, 30, 20), 30),
  indicator = "pinned", location = "nation", age = "total", sex = "total"
)
pin <- pin[reporting_date <= as.Date("2023-01-02") + 7 * 29]
tri_pin <- csfmt_reporting_triangle_v3(
  pin, id_cols = c("indicator", "location", "age", "sex")
)
reporting_completion_v1(tri_pin, max_delay_days = 15)[
  , .(n_settled, mean_delay, complete_by_md, pct_delay0, pct_delay7, pct_delay14)]
```

`pct_delay0` is 50, not 80: it is the share reported on the Monday of the
reference week, delay day 0. `pct_delay7` is 80, the cumulative share through
delay day 7. `mean_delay` is `0*0.50 + 7*0.30 + 14*0.20 = 4.9` days.

**Coming from an older script?** These columns used to be counted in WEEKS. On
the old weekly axis, `pct_delay1` meant "in by the end of the week after the
reference week", which is now `pct_delay13`. The names kept their shape and
changed their unit. So old code reads a column that exists and gets a different
quantity. Check every `pct_delayD` that you use.

`n_settled` is 28, not 30. A reference week counts only once
`as_of - reference Monday >= max_delay_days - 1` days, which excludes the two
newest of the 30. A second filter then drops a counted week with a total of
zero, and `n_settled` counts what passes both. No week here is empty, so the
age rule alone sets the number. The same holds on the working triangle:

```{r}
completion <- reporting_completion_v1(tri, max_delay_days = max_delay_days)
c(age_eligible = sum(age >= max_delay_days - 1L), reported = completion$n_settled)
```

The two agree only because this series has no week with a zero count. On an
indicator with quiet weeks, such as a rare pathogen or a small stratum,
`n_settled` is the smaller. A period with fewer than three weeks left gets no
row at all.

### A worked reference week

Take reference week **2023-07**. `cstime` gives its dates, and those of the two
weeks after it:

```{r}
cstime::dates_by_isoyearweek[
  isoyearweek %in% c("2023-07", "2023-08", "2023-09"),
  .(isoyearweek, isoyear, mon, thu, sun)
]
```

For a case with reference week 2023-07:

| statistic | covers reports up to | which is |
|---|---|---|
| `pct_delay0` | the Monday of the reference week | Monday 2023-02-13 |
| `pct_delay6` | end of ISO week 2023-07 | Sunday 2023-02-19 |
| `pct_delay13` | end of ISO week 2023-08 | Sunday 2023-02-26 |
| `pct_delay20` | end of ISO week 2023-09 | Sunday 2023-03-05 |

So `pct_delay6` covers the seven days from Monday 2023-02-13. `pct_delay13`
covers the **14** days from the same Monday, not the seven days of week 2023-08
alone. The columns are cumulative.

### The weekday of the extract

**The weekday of the extract changes the delay arithmetic.** A delay is the
number of days from the Monday of the reference week to the reporting date. So
the seven days of one reporting week land on seven different delay days:

```{r}
days <- seq(as.Date("2023-02-13"), as.Date("2023-02-19"), by = "day")
data.table(date = days, weekday = weekdays(days),
           delay_day_for_2023_07 = as.integer(days - as.Date("2023-02-13")))
```

On the old weekly axis, all seven of those days shared one label and one delay.

Nothing in the pipeline reads the system clock. The as-of boundary is the
newest reporting date *in the data*: `as_of` is `max(reporting_date)`. So a
Monday extract and a Sunday extract of the same week have different as-of
dates. The age in days of every reference week moves with them.

`nowcast_censor()` shows exactly what a Monday extract would have held. It drops
every cell reported after the given date:

```{r}
monday_as_of <- attr(tri, "as_of") - 6L         # the Monday of the as-of week
tri_mon <- nowcast_censor(tri, as_of = monday_as_of)

rbind(
  cbind(extract = "Sunday (week complete)",
        reporting_completion_v1(tri, max_delay_days = max_delay_days)[
          , .(n_settled, mean_delay, pct_delay6, pct_delay13, pct_delay20, pct_delay27)]),
  cbind(extract = "Monday (six days earlier)",
        reporting_completion_v1(tri_mon, max_delay_days = max_delay_days)[
          , .(n_settled, mean_delay, pct_delay6, pct_delay13, pct_delay20, pct_delay27)])
)
```

The pooled curve barely moves. `n_settled` moves, and that is the age test at
work. A week counts once `as_of - reference Monday >= max_delay_days - 1`, so
an `as_of` six days earlier moves the settled boundary back with it.

On this triangle, one reference week is at risk of a total that is too low: the
newest settled one. Its age is exactly `max_delay_days - 1`, so its last delay
cell falls on the extract day and is still filling. Every newer week is not
settled. Every older week finished reporting earlier, but only because
`sim_reports()` emits no delay beyond day 34. A report at delay
`max_delay_days` or later counts in the last delay column. So on real data, an
older settled week can also still be filling on the extract day. Count the
weeks that moved, and do not assume it is one:

```{r}
truth_sun <- nowcast_truth(tri, max_delay_days)
truth_mon <- nowcast_truth(tri_mon, max_delay_days)
cmp <- merge(truth_sun, truth_mon, by = "reference",
             suffixes = c("_sunday", "_monday"))
c(settled_sunday = nrow(truth_sun), settled_monday = nrow(truth_mon),
  in_both = nrow(cmp), totals_that_moved = nrow(cmp[truth_sunday != truth_monday]))
```

The reason bounds the whole effect on this triangle. The only cell at risk is
the delay-34 numerator of the newest settled week, which holds what arrived on
the extract day itself:

```{r}
newest_settled <- ref_weeks[max(which(age >= max_delay_days - 1L))]
triangle_long[isoyearweek_reference == newest_settled &
              reporting_date == attr(tri, "as_of"),
              .(isoyearweek_reference, reporting_date, numerator, denominator)]
```

Nothing arrived in that cell, so the settled total of this week cannot move.
**The weekday effect is bounded by the count in the last delay cell.** Widen
`max_delay_days` past the day where reporting really stops, and that cell
empties. A daily axis makes the cell one day wide instead of one week wide, so
there is less count in it to lose.

**The effect is small here because of scale, and that does not carry over.**
The cell at risk is the last delay day of one week, pooled with hundreds of
settled weeks. A shorter series, a shorter horizon or a heavier reporting tail
makes the same mechanism material. A period with the minimum of three weeks
gives that one cell a third of the weight.

The general statement is conditional. A mid-week extract undercounts the newest
settled week by the part of its last delay cell that has not arrived yet. It
also undercounts an older settled week by every late report that has not
arrived yet. `nowcast_truth()` scores a backtest against that total. So run the
backtest on an end-of-day extract, and record which day it was.

### "As of today", for the weeks on screen

With `as_of` = `r attr(tri, "as_of")` and `max_delay_days` = `r max_delay_days`,
the reference weeks fall into three groups:

```{r}
data.table(
  isoyearweek = ref_weeks,
  days_observed = age + 1L,
  status = fifelse(age >= max_delay_days - 1L, "settled",
           fifelse(age >= 7L, "still filling", "current week"))
)[, .N, keyby = status]
```

The weeks that are not settled are the weeks that the nowcast of stage 1
completed, and the weeks that the completion table leaves out. The completed
weeks are those whose nowcast rises above the count reported so far:

```{r}
q[numerator_nowcasted_q95x0 > original, isoyearweek]
```

### A change in reporting speed: `period = "year"`

`period = "year"` returns **one row per ISO year that qualifies**, and it shows
a reporting system that speeds up or slows down. One pooled curve cannot: it
averages the regimes and describes neither.

"Qualifies" matters. A period needs at least three settled weeks with a total
above zero. A period below that gets no row, and no warning. The year in which
your data starts or ends is the usual loss. A missing year means too few weeks,
never zero delay.

```{r}
reporting_completion_v1(tri, max_delay_days = max_delay_days, period = "year")[
  , .(period, n_settled, mean_delay, pct_delay6, pct_delay13, pct_delay20, pct_delay27)]
```

That is the change built into `sim_reports()`, found from the triangle. Read
`pct_delay6`, the share in by the end of the reference week itself. It runs
47.3, 44.7, 45.9, 44.3 and 44.6 across ISO 2019 to 2023, then steps to 70.7 in
2024. `mean_delay` steps the other way, from about 9.5 days to 6.06. A flat run
and then a step is what a real regime change looks like. A single year out of
line with its neighbours is usually noise.

The pooled `pct_delay6` of 47.9 sits just above every one of the slow years.
The slow weeks outnumber the fast ones 227 to 14. So the pooled curve is the old
regime with a little of the new one in it, not a mean of the two. It will drift
year by year as 2024 gains weeks.

The slices are by **ISO year**, not calendar year. The ISO year of a week is the
calendar year of its Thursday. That rule decides the year of a week at the turn
of the year:

```{r}
cstime::dates_by_isoyearweek[
  isoyearweek %in% c("2022-52", "2023-01"), .(isoyearweek, isoyear, mon, thu, sun)
]
```

ISO week 2022-52 has its Thursday on 2022-12-29, so it belongs to ISO year
2022, although its Sunday, 2023-01-01, is in calendar 2023. ISO week 2023-01
has its Thursday on 2023-01-05, and belongs to 2023.

`period = "month"` slices finer, and shows *when* a change happened. It uses the
same Thursday rule to give a week that spans two months to one of them:

```{r}
tail(reporting_completion_v1(tri, max_delay_days = max_delay_days, period = "month")[
  , .(period, n_settled, mean_delay, pct_delay6, pct_delay13)], 6)
```

The step falls between 2023-12 and 2024-01. Note the small `n_settled` per
month, four or five weeks, so one month alone is noisy. Read the sequence, not
one row.

### Every number here depends on `max_delay_days`

`reporting_completion_v1()` works from a triangle whose last delay column,
`max_delay_days - 1`, also holds every report at a later delay. So its
denominator is the total reported by `as_of`, late reports included. Two
results follow, whatever the real reporting tail looks like:

- `complete_by_md` is the last cumulative share of that total, so it is 1.
- The last column, `pct_delay<max_delay_days - 1>`, is that share as a
  percentage, so it is 100.

Neither shows the tail. The column before the last one does:
`100 - pct_delay<max_delay_days - 2>` is the share reported at delay day
`max_delay_days - 1` or later. `mean_delay` counts each of those reports at
delay day `max_delay_days - 1`, so it is a lower bound. The sweep below runs one
to eight weeks of horizon, in days, and reads the last column by name each
time:

```{r}
sens <- rbindlist(lapply(7L * (1:8), function(md) {
  r <- reporting_completion_v1(tri, max_delay_days = md)
  data.table(max_delay_days = md, n_settled = r$n_settled,
             mean_delay = r$mean_delay, complete_by_md = r$complete_by_md,
             last_col = paste0("pct_delay", md - 1L),
             last_pct = r[[paste0("pct_delay", md - 1L)]],
             pct_delay6 = r$pct_delay6, pct_delay13 = r$pct_delay13)
}), fill = TRUE)
sens
c(complete_by_md_always_1 = all(sens$complete_by_md == 1),
  last_pct_always_100     = all(sens$last_pct == 100))
```

The same holds in every period, where it is easiest to mistake for a finding:

```{r}
per <- rbindlist(lapply(c("all", "year", "month"), function(p) {
  r <- reporting_completion_v1(tri, max_delay_days = max_delay_days, period = p)
  data.table(period_arg = p, rows = nrow(r),
             complete_by_md_all_1 = all(r$complete_by_md == 1),
             pct_delay34_all_100 = all(r$pct_delay34 == 100))
}))
per
```

**The diagnostic that works is the `max_delay_days` sweep itself.** Read the
`sens` table above from top to bottom. `mean_delay` climbs from 4.54 days at a
7-day horizon to 9.15 at 35 days, then moves only to 9.20 at 56. That flat end
is what *supports* `max_delay_days <- 35L` here. It is a sensitivity analysis,
not a proof. `pct_delay6` is 100 at a 7-day horizon, where it is the last
column. From 14 days on it moves only from 48.2 to 47.6, because the total at
every horizon holds every report.

A `mean_delay` that kept climbing would be clear evidence that the last column
still caps a real tail. A flat end is weaker evidence the other way. A sparse
tail and a change in the mix of settled weeks both flatten the curve.

Two cautions on the sweep:

- The short horizons are biased. At a 7-day horizon every report from day 6 on
  counts at day 6. So `pct_delay6` is 100 by construction, and `mean_delay` is
  too low.
- Not all the movement past 35 days comes from the tail. `n_settled` falls down
  the rows, because a longer horizon settles fewer weeks. The weeks it drops are
  the newest, which report fast on this series. That lowers the pooled
  `pct_delay6` a little, on composition alone. Compare rows with equal
  `n_settled`, or read `period = "year"`, before you call a small drift a tail.

`sim_reports()` emits no delay beyond day 34. So on *this* triangle, a horizon
of 35 days puts no later report in the last column, and the flat end is exact.
We know that only because we can read the generator. On a real series you
cannot check it. Widen the horizon until `mean_delay` stops moving. The share
reported on the last delay day or later is then
`100 - pct_delay<max_delay_days - 2>`. Only a still wider horizon shows how that
share spreads over the later days.

## 4. Rate: a nowcast numerator over a nowcast denominator

From here on, every stage takes the ensemble and returns the ensemble.

**Estimand.** The percentage of tests that were positive, per reference week.
It is computed **per draw**, so it carries the uncertainty of both the
numerator and the denominator. A division of the collapsed medians would lose
that, and give a ratio that no draw produced.

Column `j` is the same Monte-Carlo draw in every measure, so the division is
element by element:

```{r}
ens <- ens_add_rate(ens,
                    numerator   = "numerator_nowcasted",
                    denominator = "denominator_nowcasted",
                    per         = 100)
names(ens$draws)
```

The naming grammar builds the name of the new measure, so build it the same way
and do not type it out:

```{r}
rate <- csfmt_var("numerator_nowcasted", denom = "denominator_nowcasted", per = 100)
rate
```

```{r}
qr <- ens_collapse(ens, probs = c(0.05, 0.5, 0.95))
tail(qr[, .(isoyearweek,
            positives = numerator_nowcasted_q50x0,
            tests     = denominator_nowcasted_q50x0,
            pct_lo    = round(get(csfmt_var(rate, q = 0.05)), 2),
            pct       = round(get(csfmt_var(rate, q = 0.50)), 2),
            pct_hi    = round(get(csfmt_var(rate, q = 0.95)), 2))], 6)
```

Two guards apply. A zero denominator gives `NA`, not a false 0% that would read
as a real drop. The numerator is a subset of the denominator, so the rate is
capped at `per`, and a draw above the cap gives a warning.

No warning appeared above, so no draw was above the cap. But the engine
nowcasts the numerator and the denominator **independently**, and nothing keeps
them coherent. On a series where the two are close, expect that warning.

## 5. Short-term trend on the rate

**Estimand.** The OLS slope of the *completed* series over the last
`trend_isoyearweeks` weeks, and that slope as a percentage of the current level.
It is **descriptive**: a summary of the six numbers in the window, not an
estimate of a latent growth parameter. Per draw, it carries the uncertainty that
the nowcast put into those six numbers.

Fit the trend to the nowcast, not to the reported counts. That is the point of
the pipeline. The reported counts turn down at the right-hand edge only because
the reports have not arrived. A trend fitted to them shows a fall that comes
from the reporting.

The method returns a slope, a growth rate and the share of draws with a
positive slope. It does **not** return an increasing or not-increasing label:
choose your own cut-off on that share. The `cstidy::csfmt_rts_data_v1` method
of `short_term_trend()` does return a status factor, from a different
estimator, as the end of this section explains.

The ensemble method fits a closed-form OLS line to the last
`trend_isoyearweeks` nowcast values, separately in every draw column. It adds
two draw matrices and one column in `$data`:

```{r}
ens_trend <- short_term_trend(ens, measure = rate, trend_isoyearweeks = 6)
setdiff(names(ens_trend$draws), names(ens$draws))
grep("_trend_", names(ens_trend$data), value = TRUE)
```

- `..._trend_beta1`: the OLS slope, in percentage points per week, per draw.
- `..._trend_gr`: the growth rate `100 * slope / level`, per draw, in percent of
  the current level per week.
- `..._trend_increasing_pr`: the share of draws with a positive slope. It is
  already a summary over the draws, so it goes in `$data`, and the collapse
  leaves it as it is. The method adds it by reference, so `ens$data` has it
  too.

Collapse as usual:

```{r}
gr <- csfmt_var(rate, role = "trend", suffix = "_gr")
qt <- ens_collapse(ens_trend, probs = c(0.05, 0.5, 0.95))
trend <- qt[, .(isoyearweek,
                gr    = round(get(csfmt_var(gr, q = 0.50)), 2),
                gr_lo = round(get(csfmt_var(gr, q = 0.05)), 2),
                gr_hi = round(get(csfmt_var(gr, q = 0.95)), 2),
                p_increasing = get(csfmt_var(rate, role = "trend",
                                             suffix = "_increasing_pr")))]
tail(trend, 10)
```

The last four rows are the nowcast weeks, and every one of their point
estimates is negative. The bands widen toward the right-hand edge. Three of the
four lie entirely below zero, and the newest runs from -31.17 to 5.14, with
`p_increasing` at 0.138. The series is past its winter peak, on the way into
spring. Only in the newest week is the nowcast uncertainty wide enough to allow
a rise.

That result says the *completed* rate fell over each six-week window, given
this nowcast. It does not say that the fall will continue. It depends on the
nowcast being right about the four weeks that are still filling. Stage 2
measured those weeks as the weakest point of the engine.

### Every week gets an interval, the settled weeks too

Two things are uncertain, and both reach the draws.

The first is the **level**. A week still filling has a nowcast, so its
numerator and denominator differ from draw to draw. A settled week does not:
every draw of its observed total is the same number.

The second is the **line** fitted through those levels. The line is never
certain, not even when the six weekly rates are known exactly. Six known points
do not fix the slope of the process that made them. So `short_term_trend()`
always adds the sampling error of the slope, `se * reference`, to each draw
before it forms the growth rate.

The slope error cannot be turned off. `p_increasing` is the share of draws with
a positive slope. Without the slope error, a settled week has the same slope in
every draw, so the share is exactly 0 or 1 for *every* settled week. A
threshold on it would then do nothing.

```{r}
settled_weeks <- truth$reference
tr_settled <- trend[isoyearweek %in% settled_weeks & !is.na(gr_lo)]
p <- tr_settled$p_increasing
data.table(
  settled_rows_with_a_full_window = nrow(tr_settled),
  any_degenerate_interval = any(tr_settled$gr_lo == tr_settled$gr_hi),
  p_increasing_strictly_inside_0_1 = sum(p > 0 & p < 1),
  p_increasing_at_a_boundary = sum(p %in% c(0, 1))
)
```

Every settled week has a real interval, and most have a `p_increasing`
strictly between 0 and 1. The rest sit at exactly 0 or 1, for a different
reason. Their windows have a slope so many standard errors from zero that every
draw falls on the same side. That is a probability of 1 at the Monte-Carlo
resolution that the number of draws buys, not a sign. Raise `n_sim` to tell
0.999 from 1.

### What that interval is, and what it is not

The slope error answers a **model-based** question. *Treat the six weekly
values as noisy observations around a latent straight line, and report the
uncertainty of its slope.* A trend arrow asks that question. It commits you to
three assumptions:

1. the weekly mean is **linear** across the window,
2. the deviations are **independent** across weeks,
3. the deviations are **homoskedastic**, with the distribution that
   `error_reference` names.

None of the three holds for this series. `sim_reports()` draws tests from a
Poisson and positives from a binomial, both with a **sinusoidal** mean. So the
mean in the window is curved, and the variance of the rate changes with the
level and with the number of tests. Read the interval as the sensitivity of the
slope, not as a validated confidence interval, and say which one you mean when
you publish it.

A practical trap comes on top of the assumptions. Under
`error_reference = "auto"`, the identity and quasi-Poisson families use a t on
`trend_isoyearweeks - 2` degrees of freedom. That is 4 at the width of 6 used
here. At the default width of 3 it is 1, a Cauchy: the growth-rate quantiles
then have no finite variance and very heavy tails. Widen the window, or pass
`error_reference = "normal"`. A Wald interval uses the normal, and so does a
pipeline built on `stats::confint()`.

Width 2 is an error for those two families. Both estimate a dispersion from
`width - 2` residual degrees of freedom, so at width 2 there are none and `se`
has no value. `family = "binomial"` fixes the dispersion at 1, and accepts
width 2.

The ensemble method is not the same estimator as `short_term_trend()` on a
`cstidy::csfmt_rts_data_v1`. That one fits a quasi-Poisson log-link model over
a moving window, and returns a status factor, `increasing` or `notincreasing`,
after Benedetti (2019); see `?short_term_trend`. The ensemble method defaults to
an OLS slope, because it is a fixed linear filter that runs down 500 draw
columns at once. `family = "quasipoisson"` gives it the log link of the
deprecated method, at the same cost, because it runs the iteration as a filter
too. **The `csfmt_rts_data_v1` method belongs to the older design, and it is
deprecated.** New work SHOULD run `short_term_trend()` on the ensemble, as
above, before `ens_collapse()`.

Carry the trend ensemble forward:

```{r}
ens <- ens_trend
```

## 6. MEM intensity thresholds

**Estimand.** Which of five seasonal intensity levels the rate of this week
falls in. The thresholds come from *earlier seasons of the same series*, by the
Moving Epidemic Method. The classification is per draw, so the answer is a
distribution over the levels, not one label.

This is the first stage that needs **history**, and a lot of it.
`mem_thresholds_v1()` needs complete earlier seasons to fit the thresholds of a
season. `min_seasons` (default 2) is a hard floor. `prefer_seasons` (default 5)
is the depth below which a fit is provisional. A season needs
`min_weeks_per_season` (default 30) weeks to count as training at all. That is
why this series starts in 2019, not last year.

```{r}
ens <- mem_thresholds_v1(ens, measure = rate)
tail(ens$data[, .(isoyearweek, mem_n_seasons, mem_preepidemic,
                  mem_medium, mem_high, mem_veryhigh)], 3)
```

Read the **message**. It says that some seasons were fit on fewer than
`prefer_seasons` training seasons, and `mem_n_seasons` records how many each
one got. Thresholds fit on two seasons are worth much less than thresholds fit
on five, and this is where you see which you have.

After the collapse, the classification is a probability per level:

```{r}
qm <- ens_collapse(ens, probs = 0.5)
pcols <- grep("_status_prob_", names(qm), value = TRUE)
intensity <- qm[, c("isoyearweek", pcols), with = FALSE]
setnames(intensity, pcols, sub(".*_status_prob_", "", pcols))
tail(intensity, 6)
```

Some weeks belong to a season with no thresholds: the early seasons, which have
no earlier seasons to learn from. Every draw of those weeks is `NA`, so they get
no level:

```{r}
status <- ens$draws[[csfmt_var(rate, role = "status")]]
c(weeks = nrow(status),
  weeks_with_no_threshold = sum(apply(status, 1, function(r) all(is.na(r)))))
```

A real influenza series would also use `exclude_seasons`. It keeps unusual
seasons, such as a pandemic year or a season with a data gap, out of the
training data. An excluded season still gets thresholds; only the data that they
are fit on changes.

## 7. Signal detection: historical limits

**Estimand.** Whether the rate of this week is above what the same week looked
like in earlier years: the historical-limits method. The baseline is the mean
and standard deviation of the same week, plus and minus one week, across
`baseline_isoyears` earlier years. The limit is the 99.5th percentile of a
normal with that mean and standard deviation. The method compares every draw
with the limit, so the output is an exceedance *probability*, not a yes or no.

The default `baseline_isoyears = 5` needs five full years of history before the
first week gets a limit. This series is shorter, so it uses three, which is the
kind of trade every new indicator faces:

```{r}
ens <- signal_detection_hlm(ens, measure = rate, baseline_isoyears = 3)
qh <- ens_collapse(ens, probs = 0.5)
hcols <- grep("_hlmstatus_prob_", names(qh), value = TRUE)
signal <- qh[, c("isoyearweek", "hlm_threshold", hcols), with = FALSE]
setnames(signal, hcols, sub(".*_hlmstatus_prob_", "p_", hcols))
tail(signal, 6)
```

`p_high` is the share of draws above the limit. It describes the *nowcast*
uncertainty: how likely the completed rate is above the historical limit. It is
not a p-value, a posterior probability or a false-alarm rate. The baseline comes
from the point history with no uncertainty attached, so `hlm_threshold` is
treated as known.

The first weeks have too little history for a baseline, and every draw of them
is `NA`:

```{r}
hlm <- ens$draws[[csfmt_var(rate, role = "hlmstatus")]]
c(weeks = nrow(hlm),
  weeks_with_no_baseline = sum(apply(hlm, 1, function(r) all(is.na(r)))))
```

## 8. Collapse: the end of the chain

`ens_collapse()` reduces every draw matrix to quantile columns, and every status
matrix to a probability per level. With `heal = TRUE` it passes the result to
`cstidy::set_csfmt_rts_data_v3()`, which adds the standard calendar columns and
returns a `csfmt_rts_data_v3`.

```{r}
final <- ens_collapse(ens, probs = c(0.05, 0.5, 0.95), heal = TRUE)
class(final)
c(rows = nrow(final), cols = ncol(final))
tail(final[, .(isoyearweek, isoyear, isoweek, season, seasonweek)], 3)
```

**This is where the pipeline ends.** A `csfmt_rts_data_v3` is a format for
plots, tables and reports, not for analysis, and there is no way back. The draws
are gone, so no per-draw trend, level probability or exceedance probability can
come after this point. Every stage that needs draws runs before the collapse,
which is why stages 4 to 7 come before it.

csdb ships table validators for `csfmt_rts_data_v1` and `v2` only, so it cannot
validate a healed v3 table.

## 9. Quality control

The quality-control functions check a run and change nothing. They return a
verdict, and the caller decides what to do with it.

`qc_surveillance_data_v1()` checks the input of one indicator before the
nowcast. It checks that the input has rows and a reference column, and that its
newest week is not older than expected:

```{r}
qc_surveillance_data_v1(triangle_long)
qc_surveillance_data_v1(triangle_long, expect_latest = "2024-19")
```

`qc_week_over_week_v1()` compares this week's run with last week's run. Here
last week's run is the same pipeline on the triangle as of seven days earlier,
which `nowcast_censor()` rebuilds. Each run nowcasts the count and adds signal
detection, which gives it a status column:

```{r}
run <- function(x) {
  set.seed(3)
  e <- nowcast_delay_ecdf_v1(x, max_delay_days = max_delay_days, n_sim = 500)
  e <- signal_detection_hlm(e, measure = "numerator_nowcasted", baseline_isoyears = 3)
  ens_collapse(e, probs = 0.5)
}
this_week <- run(tri)
last_week <- run(nowcast_censor(tri, as_of = attr(tri, "as_of") - 7L))

qc <- qc_week_over_week_v1(this_week, last_week, max_delay_weeks = 5)
qc$integrity
qc$signal
```

`$integrity` lists the settled weeks whose published median changed between the
two runs. It is empty here, because no report in this series arrives after its
week settles. `$signal` lists the status changes in the newer weeks, and a new
week counts as a change.

Both tables come from `compare_results()`, which holds every value column of
every week in both runs. Here are the nowcast medians of the newer weeks, which
move from one run to the next:

```{r}
changes <- compare_results(this_week, last_week)
changes[column == "numerator_nowcasted_q50x0" & isoyearweek >= "2024-14",
        .(isoyearweek, cur = round(cur, 1), prv = round(prv, 1))]
```

Both functions need identity columns with csfmt schema names, such as
`indicator_tag` and `location_code`, as this triangle has. `?compare_results`
explains the error you get otherwise.

## The naming grammar

The pipeline builds its column names from named parts. So code can find a
column by its parts, and does not need the name written out:

```
<measure>[_vs_<denom>][_<role>][_<q-coord> | _prob_<level>][_pr<per>][<suffix>]
```

```{r}
csfmt_var("numerator", role = "nowcasted", q = 0.5)
csfmt_parse("numerator_nowcasted_q50x0")
```

`csfmt_parse()` is **not** the inverse of `csfmt_var()`. It removes parts from
the right, and matches a role against a fixed list. So it cannot tell which of
two role words was the role, and it gets the denominator of the rate name wrong:

```{r}
nm <- csfmt_var("numerator_nowcasted", denom = "denominator_nowcasted", per = 100)
nm
csfmt_parse(nm)$denom
```

The `_nowcasted` of the denominator was read as the role, so `"denominator"`
comes back instead of `"denominator_nowcasted"`. The parse is reliable for a
name with one role word, such as `numerator_nowcasted_q50x0`. Check the result
when a measure or a denominator itself ends in a role word.

`q_label()` and `q_value()` map a probability to its column label and back.
They are not an exact inverse pair either. `q_label()` holds one decimal-percent
digit and exactly two integer-percent digits. So a finer probability is
rounded, and `p = 1` gives a three-digit label that `q_value()` cannot read.

```{r}
q_value(q_label(0.0125))   # rounded to the label grid
q_label(1)
q_value(q_label(1))        # NA: three integer digits do not parse
```

## Where next

- Run the chain on several series at once. Every stage works over the
  `time_series_id` axis, so a triangle with many locations or age groups runs
  unchanged. The trend masks the windows that would cross two series, and MEM
  fits each series on its own.
- `nowcast_passthrough_to_ensemble_v1()` replaces the engine at stage 1 for an
  indicator that SHOULD NOT be nowcast. Every later stage is the same, so an
  indicator can skip the nowcast and keep the rest of the pipeline.
- `reporting_completion_trend_v1()` puts the year and month slices of stage 3
  into one table with a `scope` column.
- `nowcast_estimate_calibration_v1()` turns a long replay into an interval
  scaling factor per horizon, as a check on an engine.
- The `cstidy::csfmt_rts_data_v1` methods of `short_term_trend()` and
  `signal_detection_hlm()` are deprecated. No vignette runs them, and their help
  pages hold the only examples. `vignette("csalert", package = "csalert")`
  explains what replaced them.
