---
title: "Short term trend"
author: "Richard Aubrey White, Chi Zhang"
date: "2023-05-31"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Short term trend}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to compute short-term trend signals using `csalert::short_term_trend`. Two scenarios are covered: a single location (Norway) and multiple locations (Norwegian counties).

> **The method shown here is deprecated.** Both scenarios below run
> `short_term_trend.csfmt_rts_data_v1`, which belongs to the pre-ensemble
> architecture. It still works and emits no warning, so existing pipelines are
> undisturbed — but new work should run `short_term_trend()` on a
> `csfmt_ensemble_v3` instead, before `ens_collapse()`. See
> `vignette("nowcasting", package = "csalert")`, which runs it as stage 5 of the
> full pipeline.
>
> **The replacement is not a drop-in.** The ensemble method takes one `measure`
> naming a draws matrix rather than `numerator` / `denominator` / `prX`, and it
> returns a per-draw slope, growth rate and P(increasing) rather than a status
> factor and a doubling time. Migrating is a rewrite of the call site, and the
> numbers will not match.

The v1 method fits a quasi-Poisson log-link model over a moving window and returns a factor status column, following Benedetti (2019). The ensemble method computes a closed-form OLS slope down every Monte-Carlo draw, which propagates a nowcast's uncertainty into the trend instead of discarding it. They are different estimators, not two interfaces to one estimator.

The data used below are reported counts, not nowcasts. `remove_last_isoyearweeks` trims the most recent weeks precisely because those counts are still arriving; a nowcast is the alternative to trimming them.

```{r}
library(ggplot2)
library(data.table)
library(magrittr)
```

## Single location

### Covid-19 hospitalisation data

The dataset contains daily and weekly counts of Covid-19 hospitalisations in Norway where Covid-19 was recorded as the primary cause. It does not distinguish between age groups or sex.

This extract was pulled on 2022-05-04 and covers 2020-02-21 to 2022-05-03. The raw data in CSV and XLSX formats are available on the [GitHub repository](https://github.com/folkehelseinstituttet/surveillance_data).

### Data in `cstidy` format

The data have been pre-processed into [cstidy format](https://niphr.github.io/cstidy/articles/csfmt_rts_data_v2.html), which provides a standardised column structure including data types and missingness information.

```{r}
d_hosp <- cstidy::nor_covid19_icu_and_hospitalization_csfmt_rts_v1
d_hosp
```

### Weekly observations

The `short_term_trend` function is run on the weekly subset of the data. The `trend_isoyearweeks` argument sets the number of weeks used to estimate the trend, and `remove_last_isoyearweeks` trims the most recent weeks to account for reporting delays.

```{r}
d_hosp_weekly <- d_hosp[granularity_time=="isoyearweek"]

res <- csalert::short_term_trend(
  d_hosp_weekly,
  numerator = "hospitalization_with_covid19_as_primary_cause_n",
  trend_isoyearweeks = 6,
  remove_last_isoyearweeks = 1
)

# create the trend label
res[, hospitalization_with_covid19_as_primary_cause_trend0_41_status := factor(
  hospitalization_with_covid19_as_primary_cause_trend0_41_status,
  levels = c("training","forecast","notincreasing", "increasing"),
  labels = c("Training","Forecast","Not increasin", "Increasing")
)]

colnames(res)
```

The function adds several new columns to the original data. The example below shows the forecasted counts alongside the trend status.

```{r}
# check some columns
res[
  ,
  .(
    date,
    hospitalization_with_covid19_as_primary_cause_n,
    hospitalization_with_covid19_as_primary_cause_forecasted_n,
    hospitalization_with_covid19_as_primary_cause_trend0_41_status
  )
]
```

### Visualising trend status

Trend status can be mapped to bar fill colours, with error bars showing the prediction interval.

```{r}
q <- ggplot(
  res,
  aes(
    x = isoyearweek,
    y = hospitalization_with_covid19_as_primary_cause_forecasted_n,
    group = 1
  )
)
q <- q + geom_col(mapping = aes(fill = hospitalization_with_covid19_as_primary_cause_trend0_41_status))
q <- q + geom_errorbar(
  mapping = aes(
    ymin = hospitalization_with_covid19_as_primary_cause_forecasted_predinterval_q02x5_n,
    ymax = hospitalization_with_covid19_as_primary_cause_forecasted_predinterval_q97x5_n
  )
)
q <- q + scale_y_continuous("Weekly hospitalization with Covid-19 as primary cause", expand = c(0, 0.1))
q <- q + scale_x_discrete("Isoyearweek")
q <- q + expand_limits(y=0)
q <- q + scale_fill_brewer("6 week trend", palette = "Set1")
q
```

Point shapes can be used instead, placing a symbol above each bar to mark the trend category.

```{r}
shape_adjustment_factor <- max(res$hospitalization_with_covid19_as_primary_cause_forecasted_n)*0.01
q <- ggplot(
  res,
  aes(
    x = isoyearweek,
    y = hospitalization_with_covid19_as_primary_cause_forecasted_n,
    group = 1
  )
)
q <- q + geom_col()
q <- q + geom_point(mapping = aes(
  y = hospitalization_with_covid19_as_primary_cause_forecasted_n + shape_adjustment_factor,
  shape = hospitalization_with_covid19_as_primary_cause_trend0_41_status
))
q <- q + geom_errorbar(
  mapping = aes(
    ymin = hospitalization_with_covid19_as_primary_cause_forecasted_predinterval_q02x5_n,
    ymax = hospitalization_with_covid19_as_primary_cause_forecasted_predinterval_q97x5_n
  )
)
q <- q + scale_y_continuous("Weekly hospitalization with Covid-19 as primary cause", expand = c(0, 0.1))
q <- q + scale_x_discrete("Isoyearweek")
q <- q + expand_limits(y=0)
q <- q + scale_shape_manual("6 week trend", values = c("Increasing" = 17, "Decreasing" = 6))
q
```

## Multiple locations

When the data contain multiple locations, `short_term_trend` computes the trend for each one. Here the function is applied to weekly Covid-19 case counts at county level.

```{r}
d <- cstidy::nor_covid19_cases_by_time_location_csfmt_rts_v1[
  granularity_time == "isoyearweek" &
  granularity_geo == "county"
]

trend <- csalert::short_term_trend(
  d,
  numerator = "covid19_cases_testdate_n",
  trend_isoyearweeks = 6,
  remove_last_isoyearweeks = 1
)

print(trend)
```

### Mapping trend status

The trend status for a single week can be joined to a county map polygon dataset and plotted geographically.

```{r}
pd <- copy(csmaps::nor_county_map_b2020_split_dt)
pd[
  trend[isoyearweek == "2021-44"],
  on = c("location_code"),
  covid19_cases_testdate_trend0_41_status := covid19_cases_testdate_trend0_41_status
]

# plot map
q <- ggplot()
q <- q + geom_polygon(
  data = pd,
  mapping = aes(x = long, y = lat, group = group,fill=covid19_cases_testdate_trend0_41_status),
  color="black",
  linewidth = 0.2
)
q <- q + coord_quickmap()
q <- q + theme_void()
q <- q + labs(title="MSIS cases per 100k population for week 2021-44")
q <- q + scale_fill_brewer("Covid trends", palette = "Set1", direction = -1)
q
```
